### 2024-11-01 README.qmd
Initial commit

---
format: gfm
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  fig.retina = 3
)

library(emoji)

# manuscript_target <- targets::tar_read(manuscript)

# yaml_metadata <- rmarkdown::yaml_front_matter(here::here("manuscript", "manuscript.qmd"))
# title <- yaml_metadata$title
# abstract <- yaml_metadata$abstract
```

## `r emoji_glue(":person_feeding_baby::bat:")`: Note on "butterscotch bat
Because project titles change all the time with revisions, rewriting, and peer review, we used [{codename}](http://svmiller.com/codename/) to generate an [Ubuntu-style](https://wiki.ubuntu.com/DevelopmentCodeNames) internal-to-us project name that won’t change.

```{r generate-code-name, eval=FALSE}
library(codename)
codename_message()
#> code name generated by {codename} v.0.5.0. R version 4.4.1 (2024-06-14).

codename(seed = "handbook donors", type = "ubuntu")
#> [1] "butterscotch bat"
```

## Licenses

**Text and figures:** All prose and images are licensed under Creative Commons ([CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)).

**Code:** All code is licensed under the [MIT License](LICENSE.md).

## Contributions and Code of Conduct

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/0/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.

### 2024-11-01 manuscript.qmd
Finish foreign aid donors section

---
title: "Donor Responses to Civil Society Repression"
subtitle: "Chapter for the *Handbook of Political Control*"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
  attributes:
    corresponding: true
bibliography: references.bib
toc: true
format:
  wordcount-html: 
    wordcount-banner: true
params:
  wordcount: |
    <strong>{{< words-sum body-note-ref >}} total words</strong>: {{< words-body >}} in the body • {{< words-ref >}} in the references • {{< words-note >}} in the notes
---

::: {.content-visible when-meta="wordcount-banner"}
{{< include _extensions/andrewheiss/wordcount/banner.html >}}
:::

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE, include = FALSE, warning = FALSE, message = FALSE,
  fig.width = 6, fig.height = (6 * 0.618), fig.retina = 3,
  out.width = "100%", fig.align = "center"
)
```

```{r}
#| label: libraries-data
library(tidyverse)
library(tidygraph)
library(ggraph)
```

Given the exponential increase in administrative and legal repression of CSOs across the globe and its effectiveness at muzzling groups, how have donors responded? How do these responses differ across foreign government and private donors? How have local philanthropists and NGOs adapted and responded to such repression? What challenges do civil society groups continue to face despite these responses?

## Foreign aid donors

A rich literature on the purposes of foreign aid explores a broad collection of reasons why states transfer money and resources to others. Aid agencies and international organizations face competing interests when deciding how to allocate resources, which projects to undertake, and which countries to work with. Donor agencies rhetorically declare normative justifications for engaging in development, with recipient-focused development goals like poverty reduction, democratization, human rights promotion, and environmental protection [@HoefflerOutram:2011; @Kallman:2020]. In practice, though, aid is given for far more instrumental reasons based on donor countries' domestic and foreign policy–oriented strategic goals [@AlesinaDollar:2000; @Apodaca:2017; @Schuller:2012], including being responsive to popular democratic pressure [@HeinrichKobayashiLong:2018], insulating themselves from regional instability [@Bermeo:2017], expanding export markets [@BarthelNeumayerNunnenkamp:2014; @Mitchell:2002], influencing recipient states' domestic policies [@Apodaca:2013], and projecting donor country values [@MorgensternBrown:2022].

### Donor responses and conditionality

A key enforcement mechanism to ensure that recipient-country actions align with donor-country goals—and thus reduce the moral hazard inherent in donating resources—is to condition aid on the achievement of specific democratizing and economic goals [@Winters:2024; @CheesemanSwedlundOBrien-Udry:2024; @HoefflerJustino:2024; @Kilby:2024; @Koch:2015]. Many donor governments condition aid on human rights outcomes, using progress in respect for human rights as leverage to shape recipient-country behavior. Conditional aid sends dual signals: it serves as an instrumental "stick" to pressure governments to change policies and an expressive "flag" to signal donor commitment to human rights norms [@DasandiErez:2023]. However, this dual-purpose strategy often creates dilemmas—if aid is withdrawn, it may jeopardize humanitarian programs or worsen political conditions. Conversely, continuing aid might signal tolerance for rights violations, weakening the credibility of the donor’s human rights agenda.

The often-conflicting trends of both pursuing normative development goals and furthering foreign policy interests leads to mixed responses when recipient countries fail to adhere to aid conditions and violate human rights. Prior research finds conflicting relationships between human rights abuses and foreign aid, variously concluding that when confronted by recipient states that violate human rights, donor agencies punish repressive states by reducing, suspending, or withdrawing aid [@LebovicVoeten:2009], increasing aid (or withdrawing less aid) to repressive states as a way to counter state violence [@Demirel-PeggMoskowitz:2009; @Nielsen:2013; @HeinrichKobayashiLong:2018; @EsareyDeMeritt:2017], or make no measurable change in aid allocations [@Neumayer:2003; @Neumayer:2003b].

One reason for such mixed results is that the decision to withdraw or suspend aid in response to human rights abuses is often seen as a binary choice driven by self-interested motivations: a donor country that sees human rights abuses in a recipient state will consider its own foreign policy interests before deciding to either continue or curtail aid. In reality, however, the decision is far more complex and involves a wide range of possible motivations and responses. @CheesemanSwedlundOBrien-Udry:2024 offer a stylized decision tree of possible choices following recipient-country human rights violations—we provide a simplified version of their framework in @fig-agency-response-tree.

```{r}
#| label: fig-agency-response-tree
#| fig-cap: Possible donor- and recipient-country responses to human rights violations
#| include: true
#| fig-width: 6
#| fig-height: 3.8
#| out-width: 80%

nodes <- tribble(
  ~id, ~x, ~y,  ~step,
  1,   0,  2.5, "Violate human rights",
  2,   1,  4,   "Do nothing",
  3,   1,  3,   "Threaten",
  4,   1,  2,   "Withdraw\nor suspend",
  5,   1,  1,   "Shift aid",
  6,   2,  4,   "Do nothing",
  7,   2,  2.5, "Modify to improve\nrelationship",
  8,   2,  1,   "Pressure donors\nto back down"
)

edges <- tibble(
  from = c(1, 1, 1, 1, 2, 2, 2, 3, 3, 3, 4, 4, 4, 5, 5, 5),
  to = c(2, 3, 4, 5, 6, 7, 8, 6, 7, 8, 6, 7, 8, 6, 7, 8)
)

actors <- tribble(
  ~actor, ~label, ~x,
  "Recipient", "Recipient government\naction", 0,
  "Donor", "Donor government\nresponse", 1,
  "Recipient", "Recipient government\nresponse", 2
) |>
  mutate(
    xmin = x - 0.5,
    xmax = x + 0.5
  )

tidy_graph <- tbl_graph(nodes = nodes, edges = edges) |> 
  create_layout(layout = nodes)

ggraph(tidy_graph) +
  geom_rect(
    data = actors, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = 4.2, fill = actor), inherit.aes = FALSE,
    alpha = 0.25
  ) +
  geom_rect(
    data = actors, aes(xmin = xmin, xmax = xmax, ymin = 4.2, ymax = Inf, fill = actor), inherit.aes = FALSE,
    alpha = 0.5
  ) +
  geom_text(
    data = actors, aes(x = x, y = 4.4, label = label), inherit.aes = FALSE,
    size = 3.5,
    fontface = "bold", lineheight = 1
  ) +
  geom_node_label(
    aes(x = x, y = y, label = step),
    label.r = unit(0, units = "pt"),
    size = 3.5, lineheight = 1, fontface = "bold"
  ) +
  geom_edge_link(
    aes(
      start_cap = label_rect(node1.step, padding = margin(5, 5, 5, 5, "pt")),
      end_cap = label_rect(node2.step, padding = margin(5, 5, 5, 5, "pt"))
    ),
    arrow = arrow(angle = 30, length = unit(6, "pt"), type = "closed"),
    edge_linewidth = 0.25, color = "grey40"
  ) +
  scale_x_continuous(expand = expansion(0)) +
  scale_fill_manual(values = c("grey40", "grey80"), guide = "none") +
  expand_limits(y = 0.9) +
  theme_void()
```

Donor countries have a range of possible responses to recipient repression, including doing nothing and ignoring the violation, threatening to change aid allocations, withdrawing or suspending aid, or shifting aid to other projects or programs. In turn, recipient governments can then respond by doing nothing and ignoring the threat or suspension, change their behavior and cease the repression to improve their relationship with the donor country, or pressure the donor country to back down and withdraw their complaint. In this simplified decision tree, there are 12 possible combinations of outcomes,^[@CheesemanSwedlundOBrien-Udry:2024 describe additional stages in their tree, with donors responding to recipients' responses, creating dozens of possible pathways.] and each pathway is conditional on a host of donor-country and recipient-country characteristics and contexts, including geopolitical dynamics in the recipient country's region, domestic political institutions within the donor country, public opinion in the donor country, coordination between coalitions of donors, the type of aid, and the type of human rights violation, among others [@CheesemanSwedlundOBrien-Udry:2024]. Such complexity makes it difficult to find consistent treatment effects in research that explores the overall effect of repression on aid—that is, we cannot definitively say that human rights causes an overall reduction or increase in aid. Instead, newer research looks at donor responses in more specific circumstances, exploring the factors that lead to specific response pathways [@Corwin:2023].

### Donor responses to civil society laws

Research on aid conditionality and agency responses to human rights abuses tends to look at more overt and violent forms of political oppression like respect for personal integrity, civil, and political rights [@Corwin:2023; @Neumayer:2003; @CheesemanSwedlundOBrien-Udry:2024], including arresting political activists, closing media outlets, suppressing protests, engaging in extrajudicial killings, and so on. The effects of more subtle administrative crackdown on aid, including the passage of anti-NGO laws and the constricting of civic space, are less explored and understood.

Civil society plays a crucial role in the distribution and implementation of international aid. Governmental aid agencies, such as the United States Agency for International Development (USAID) and the United Kingdom's Foreign, Commonwealth and Development Office (FCDO), channel substantial funding through NGOs in recipient countries. TODO: Something about the amount of money that goes through CS, perhaps highlighting just the US or UK. Larger international organizations also rely on domestic NGOs to carry out humanitarian and development projects. TODO: Something about the amount of money that the World Bank or UNHCR sends to NGOs. Outsourcing the provision of aid to NGOs allows donors to avoid aid capture in weak institutional environments and can lead to more effective program implementation [@Dietrich:2013; @CruzLabonneWright:2024]. Given the importance of NGOs in aid delivery, we have reason to expect that donor responses to civil society crackdown abroad are treated differently than more violent forms of repression, and thus likely to follow different pathways in @fig-agency-response-tree.

Donor countries have a range of possible responses when recipient countries engage in administrative crackdown. Recent research finds that national and multilateral aid agencies are responsive to anti-NGO laws, though often in inconsistent ways [@ChaudhryHeiss:2022a]. @ChristensenWeinstein:2013 show that aid from bilateral donors—or direct country-to-country aid—sees an average decrease of $25–\$50 million after recipient states pass laws that restrict the flow of foreign funds to NGOs. However, aid from international, multilateral agencies is not affected by these laws. Similarly @DupuyPrakash:2018 find that the adoption of a restrictive foreign funding law is associated with a 32% decline in aid from national, bilateral donors, but has no corresponding change in aid from international, multilateral donors.

One key reason for the divergence in the responses of bilateral and multilateral donors is the degree to which these donors rely on recipient-country NGOs to implement programs. Donor governments and agencies have different experiences with NGOs and correspondingly prioritize the openness of civil society in recipient countries in different ways. Multilateral donors like the World Bank and OECD tend to channel money directly to recipient governments rather than domestic NGOs and are thus more insulated from the effects of local civil society repression, which likely explains the lack of formal response to anti-NGO laws [@ChristensenWeinstein:2013; @DupuyPrakash:2018]. There is more variation, however, in how bilateral donors work with domestic NGOs. For example, @AllenFlynn:2018 find that the domestic political environments of donor countries shape how those countries channel their aid abroad. Because they tend to focus more on poverty alleviation initiatives, states with left-leaning governments typically channel more aid through recipient-country NGOs. In contrast, right-leaning governments are typically interested in promoting the economic and political interests of their own states, and accordingly prefer to channel aid directly through recipient-country governments [@AllenFlynn:2018]. Human rights abuses in recipient countries will shape the responses of donor countries depending on the saliency of the threat to aid provision: political repression that threatens business interests and investments will be more salient to right-leaning governments, while anti-NGO civil society administrative crackdown will likely have a stronger effect on left-leaning governments' responses, since domestic NGOs are central to donor countries' aid distribution strategies.

The nature of civil society restrictions also influences how official donor agencies respond. In their initial work on anti-NGO restrictions, @ChristensenWeinstein:2013 cataloged civil society laws into three broader categories—barriers to funding, barriers to entry, and barriers to advocacy—but they, and others like @DupuyPrakash:2018, dedicated the bulk of their analysis to looking at the effects of barriers to funding. This focus on funding restrictions is warranted, as most authoritarian civil society legal restrictions have been aimed specifically at reducing or controlling foreign funding [@Heiss:2019; @Carothers:2015a; @CarothersBrechenmacher:2014]. Moreover, the relationship between foreign aid and foreign funding restrictions appears to be bidirectional, as states are more likely to pass restrictive foreign funding laws in response to increases in bilateral aid [@DupuyRonPrakash:2016], especially if that aid arrives during competitive elections or times of political protest or instability [@Chaudhry:2022; @Heiss:2017]. Foreign funding restrictions and aid are thus linked: increasing aid leads to more funding restrictions [@DupuyRonPrakash:2016], which in turn lead to reductions in aid [@DupuyPrakash:2018].

Though the majority of anti-NGO laws have targeted foreign funding, other laws seek to impose barriers to entry, such as strict and burdensome registration requirements, and barriers to advocacy that explicitly limit NGO political participation [@Chaudhry:2022]. Donor countries respond differently to these types of oppression. @ChaudhryHeiss:2024 (see also @ChaudhryHeiss:2022a) find that bilateral donors are quite responsive to barriers to advocacy—on average, total foreign aid decreases by 30–40% (or \$100 in a typical country) following the passage of an advocacy-focused NGO law. @RightSpringmanWibbels:2024 find similar and more nuanced trends in donor responses: donors committed to political advocacy   (i.e. those that fund democracy and civil society promotion activities) reduce funding for advocacy programs by more than 70% in response to new restrictions, and the reduction in aid persists for several years.

Overall bilateral aid does not decrease significantly in response to new barriers to entry, but the distribution channels of that aid shift as donor countries channel more aid through domestic rather than foreign NGOs [@ChaudhryHeiss:2024]. USAID even explicitly encourages its program managers to rely more on domestic, in-country NGOs when facing a restrictive legal environment [@USAID:2014, 6–7]. In addition to changing the amounts and recipients of foreign aid, donor agencies often adjust the focus of the programs they fund, redirecting funding away from programs related to politically sensitive issues like human rights, anti-corruption, and democratic development and towards tamer topics that are more compatible with recipient-country preferences, such as education, health, and humanitarian services [@Bush:2015; @CheesemanDesrosiers:2023].

These shifts in funding patterns in response to restrictive civil society laws have had measurable effects on NGOs in recipient countries. Samantha Power, current administrator of USAID, has pledged to send 25% of USAID funding to local NGOs over the next few years, though practical obstacles remain. New local partners take longer to vet and train compared to established INGO partners. Turning to local organizations entails greater risk, as domestic NGOs can be easily shut down based on the whims of the government. Additionally,  communities often distrust local groups that accept Western aid, especially those focused on democracy assistance. Donor preferences have also changed. Over the past two decades, bilateral donors have increasingly provided short-term grants for projects that have easily measurable, quantifiable outputs [@Bush:2015]. This model may work well for civic groups working on relatively tame causes or exclusively development-focused goals, but more contentious organizations facing government crackdown may need flexible funding that does not bring immediate results or does not have easily measurable goals.

## Private donors and foundations

- Why individual donors are important
- Donor reactions are more than just binary, but there now fancy figure 1 decision tree yet, since (1) the existing literature is much smaller, and (2) individual donors have less power in relation to states. Countries don’t counter-react to donor shifts in behavior
- Different pathways exist, though, depending on donor preferences and organization characteristics. As with diverging studies in the effect of repression on formal aid, there are different effects. Experiment paper + simulation paper + conjoint paper (getting at donor and org features)

In other cases, donors shift their aid to different programs. - both good and bad. Bush and taming + USAID

While research in public policy examines philanthropic giving to NGOs, the discipline of political science and international relations tends to see philanthropy as less relevant. However, philanthropy can help us understand changes in the dynamics of global governance. Private individuals exercise power and influence in unique ways that differ from other non-state actors. Understanding the dynamics of supporting international causes is important as philanthropy towards such causes can be harder to motivate because the number of recipients is larger and further removed from the donor.

While private sources can help raise such funds, official donor aid may not be easily replaceable by other sources. Testimony at a US Senate Committee on Foreign Relations hearing maintained that, “Neither remittances nor philanthropy can replace the expertise, scale, or agenda setting capacity of American foreign policy assistance.” Thus, despite promising new research on how to raise funds from private sources for besieged NGOs abroad, official aid donors also need to mount a coherent response to the shrinking civic space.

### 2024-11-01 README.qmd
Fix codename emoji

@@
 
-## `r emoji_glue(":person_feeding_baby::bat:")`: Note on "butterscotch bat
+## `r emoji_glue(":candy::bat:")`: Note on "butterscotch bat
 Because project titles change all the time with revisions, rewriting, and peer review, we used [{codename}](http://svmiller.com/codename/) to generate an [Ubuntu-style](https://wiki.ubuntu.com/DevelopmentCodeNames) internal-to-us project name that won’t change.

### 2024-11-01 README.qmd
oof fix project name heading

@@
 
-## `r emoji_glue(":candy::bat:")`: Note on "butterscotch bat
+## `r emoji_glue(":candy::bat:")`: Note on "butterscotch bat" project name
+
 Because project titles change all the time with revisions, rewriting, and peer review, we used [{codename}](http://svmiller.com/codename/) to generate an [Ubuntu-style](https://wiki.ubuntu.com/DevelopmentCodeNames) internal-to-us project name that won’t change.

### 2024-11-01 manuscript.qmd
Super minor edits

@@
 Donor countries respond differently to these types of oppression.
-@ChaudhryHeiss:2024 (see also @ChaudhryHeiss:2022a) find that bilateral donors are quite responsive to barriers to advocacy—on average, total foreign aid decreases by 30–40% (or \$100 in a typical country) following the passage of an advocacy-focused NGO law.
+@ChaudhryHeiss:2024 find that bilateral donors are quite responsive to barriers to advocacy—on average, total foreign aid decreases by 30–40% (or \$100 in a typical country) following the passage of an advocacy-focused NGO law [see also @ChaudhryHeiss:2022a].
 @RightSpringmanWibbels:2024 find similar and more nuanced trends in donor responses: donors committed to political advocacy   (i.e.
@@
 Overall bilateral aid does not decrease significantly in response to new barriers to entry, but the distribution channels of that aid shift as donor countries channel more aid through domestic rather than foreign NGOs [@ChaudhryHeiss:2024].
-USAID even explicitly encourages its program managers to rely more on domestic, in-country NGOs when facing a restrictive legal environment [@USAID:2014, 6–7].
+USAID even explicitly encourages its program managers to rely more on domestic, in-country NGOs when facing a restrictive legal environment [@USAID:2014].
 In addition to changing the amounts and recipients of foreign aid, donor agencies often adjust the focus of the programs they fund, redirecting funding away from programs related to politically sensitive issues like human rights, anti-corruption, and democratic development and towards tamer topics that are more compatible with recipient-country preferences, such as education, health, and humanitarian services [@Bush:2015; @CheesemanDesrosiers:2023].

### 2024-11-02 manuscript.qmd
Add private donor response section

@@
 bibliography: references.bib
+csl: csl/american-sociological-association.csl
 toc: true
@@
 These shifts in funding patterns in response to restrictive civil society laws have had measurable effects on NGOs in recipient countries.
-Samantha Power, current administrator of USAID, has pledged to send 25% of USAID funding to local NGOs over the next few years, though practical obstacles remain.
+Samantha Power, current administrator of USAID, has pledged to send 25% of USAID funding to local NGOs over the next few years [@Power:2021], though practical obstacles remain.
 New local partners take longer to vet and train compared to established INGO partners.
 Turning to local organizations entails greater risk, as domestic NGOs can be easily shut down based on the whims of the government.
-Additionally,  communities often distrust local groups that accept Western aid, especially those focused on democracy assistance.
+Additionally,  communities often distrust local groups that accept Western aid, especially those focused on democracy assistance [@Barkan:2012].
 Donor preferences have also changed.
 Over the past two decades, bilateral donors have increasingly provided short-term grants for projects that have easily measurable, quantifiable outputs [@Bush:2015].
-This model may work well for civic groups working on relatively tame causes or exclusively development-focused goals, but more contentious organizations facing government crackdown may need flexible funding that does not bring immediate results or does not have easily measurable goals.
+This model may work well for civic groups working on relatively tame causes or exclusively development-focused goals, but more contentious organizations facing government crackdown may need flexible funding that does not bring immediate results or does not have easily measurable goals [@Herrold:2020].
 
-## Private donors and foundations
+## Private donors
 
-- Why individual donors are important
-- Donor reactions are more than just binary, but there now fancy figure 1 decision tree yet, since (1) the existing literature is much smaller, and (2) individual donors have less power in relation to states.
-Countries don’t counter-react to donor shifts in behavior
-- Different pathways exist, though, depending on donor preferences and organization characteristics.
-As with diverging studies in the effect of repression on formal aid, there are different effects.
-Experiment paper + simulation paper + conjoint paper (getting at donor and org features)
+As explored above, given the centrality of aid in domestic foreign policy decisions and economic development strategy, substantial research in development economics, political science, and international relations has looked at the motivations behind bilateral and multilateral foreign aid, especially in response to human rights abuses in recipient states.
+However, states and international organizations are not the only institutions that engage in international giving.
+Private individuals exercise power and influence in unique ways that differ from other non-state actors.
+Since 2000, foreign aid funding from philanthropic foundations and individual donors in the United States has more than doubled [@DesaiKharas:2018], with nearly \$30 billion donated to international affairs and development organizations in 2023 [@GivingUSA:2024].
+This growth in private foreign aid has had a substantial effect on overall international  aid flows and often serves as a complement to official bilateral and multilateral aid, particularly in response to acute humanitarian crises [@DesaiKharas:2018].
 
-In other cases, donors shift their aid to different programs. - both good and bad. Bush and taming + USAID
+Understanding the dynamics of supporting international causes is important as philanthropy towards such causes can be harder to motivate.
+Work in nonprofit and philanthropic studies has long explored the personal motivations of individual donors, including reputation, altruism, and psychological benefits [see @BekkersWiepking:2011 for an exploration of eight different motivating mechanisms], but less research—particularly in political science and international  relations—has been done regarding the organizational or institutional mechanisms that drive donor behavior.
+Donors are motivated by the broader political and institutional contexts of NGO programming, and individuals are less likely to donate to foreign nonprofits and prefer giving to local causes, since the result of their giving is more visible [@CasaleBaumann:2015; @Tremblay-BoirePrakash:2017; @Wiepking:2010].
 
-While research in public policy examines philanthropic giving to NGOs, the discipline of political science and international relations tends to see philanthropy as less relevant.
-However, philanthropy can help us understand changes in the dynamics of global governance.
-Private individuals exercise power and influence in unique ways that differ from other non-state actors.
-Understanding the dynamics of supporting international causes is important as philanthropy towards such causes can be harder to motivate because the number of recipients is larger and further removed from the donor.
+Donor motivations influence how individuals respond when potential recipient organizations face legal trouble abroad.
+However, the exact menu of responses is more limited than what is available to bilateral and multilateral states, as seen previously in @fig-individual-response-tree, and far more understudied.
+@fig-individual-response-tree presents a preliminary framework for outlining possible individual donor responses to anti-NGO repression.
+When a potential recipient NGO faces legal crackdown in its host country, thus limiting its operations, interested individual donors can respond by (1) increasing their donations, as a sign of support and solidarity, (2) decreasing their donations, punishing the NGO for doing something to run afoul of its host government, or (3) not considering the host country legal environment and making no change in their donation behavior.
+@ChaudhryDotsonHeiss:2021 explore how different combinations of donor characteristics determine which response individuals are likely to take when NGOs face crackdown.
+They find that donors with longer experiences with the nonprofit sector and high levels of social trust—i.e.
+those who frequently volunteer, regularly donate to charity, and trust political institutions—are more likely to maintain their support for international NGOs that face criticism or crackdown abroad.
 
-While private sources can help raise such funds, official donor aid may not be easily replaceable by other sources.
-Testimony at a US Senate Committee on Foreign Relations hearing maintained that, “Neither remittances nor philanthropy can replace the expertise, scale, or agenda setting capacity of American foreign policy assistance.” Thus, despite promising new research on how to raise funds from private sources for besieged NGOs abroad, official aid donors also need to mount a coherent response to the shrinking civic space.
+```{r}
+#| label: fig-individual-response-tree
+#| fig-cap: Possible individual responses to civil society restrictions in recipient countries
+#| include: true
+#| fig-width: 6
+#| fig-height: 2.8
+#| out-width: 80%
+
+nodes <- tribble(
+  ~id, ~x, ~y, ~step,
+  1,   0,  3,  "Restrict civil society",
+  2,   1,  4,  "Do nothing",
+  3,   1,  3,  "Increase\ndonations",
+  4,   1,  2,  "Decrease\ndonations",
+  5,   2,  3,  "?"
+)
+
+edges <- tibble(
+  from = c(1, 1, 1, 2, 3, 4),
+  to =   c(2, 3, 4, 5, 5, 5)
+)
+
+actors <- tribble(
+  ~actor, ~label, ~x,
+  "Recipient", "NGO host government\naction", 0,
+  "Donor", "Individual donor\nresponse", 1,
+  "NGO", "NGO\nresponse", 2
+) |>
+  mutate(
+    xmin = x - 0.5,
+    xmax = x + 0.5
+  )
+
+tidy_graph <- tbl_graph(nodes = nodes, edges = edges) |> 
+  create_layout(layout = nodes)
+
+ggraph(tidy_graph) +
+  geom_rect(
+    data = actors, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = 4.2, fill = actor), inherit.aes = FALSE,
+    alpha = 0.25
+  ) +
+  geom_rect(
+    data = actors, aes(xmin = xmin, xmax = xmax, ymin = 4.2, ymax = Inf, fill = actor), inherit.aes = FALSE,
+    alpha = 0.5
+  ) +
+  geom_text(
+    data = actors, aes(x = x, y = 4.4, label = label), inherit.aes = FALSE,
+    size = 3.5,
+    fontface = "bold", lineheight = 1
+  ) +
+  geom_node_label(
+    aes(x = x, y = y, label = step),
+    label.r = unit(0, units = "pt"),
+    size = 3.5, lineheight = 1, fontface = "bold"
+  ) +
+  geom_edge_link(
+    aes(
+      start_cap = label_rect(node1.step, padding = margin(5, 5, 5, 5, "pt")),
+      end_cap = label_rect(node2.step, padding = margin(5, 5, 5, 5, "pt"))
+    ),
+    arrow = arrow(angle = 30, length = unit(6, "pt"), type = "closed"),
+    edge_linewidth = 0.25, color = "grey40"
+  ) +
+  scale_x_continuous(expand = expansion(0)) +
+  scale_fill_manual(values = c("grey40", "grey5", "grey80"), guide = "none") +
+  expand_limits(y = c(1.8, 4.5)) +
+  theme_void()
+```
+
+Additional experimental evidence examines how donors react to the knowledge that potential recipient NGOs face legal difficulties.
+In a vignette-based factorial experiment, @ChaudhryHeiss:2021 find that foreign civil society restrictions serve as a heuristic when deciding to donate—learning about NGO repression increases individual generosity and causes potential donors to be willing to give substantially more to legally restricted nonprofits.
+This effect is strongest for privately-funded human rights-focused NGOs that face legal crackdown, as individuals feel that the organization deserves their support.
+Several participants justified their increased support because the experimental organization is "doing good work in countries where it is tough for groups like them to operate and they
+need all the help they can get" [@ChaudhryHeiss:2021, 496].
+Building on this work, @ChaudhryDotsonHeiss:2024 use a survey experiment to examine how organizational characteristics (e.g., issue area, funding sources) and practices (e.g., financial transparency) interact with structural concerns like legal crackdown abroad.
+They find that, all else equal, legal crackdown on its own causes donors to reduce their support for nonprofits—individuals are 5–10 percentage points less likely to donate to an NGO that is criticized or restricted, compared to an organization that enjoys a friendly relationship with its host government.
+However, this reduction in support can be offset by organizational characteristics.
+Financial transparency and accountability protect each increase the probability of donations by nine percentage points under the worst conditions of legal crackdown.
+Individual donors thus typically reduce their support when seeing legal crackdown, but change their response and increase their when seeing other signs that the organization follows best practices and is more deserving.
+
+The question of how organizations respond to different donor reactions to legal crackdown remains unexplored (see the far right panel of @fig-individual-response-tree).
+To our knowledge, no work has been done regarding individual-level nonprofit fundraising strategies in the face of closing civic space, opening promising avenues for future research.
+
+While private donations have become an important source of funding in the age of closing civi space and increased restrictions on official foreign funding, they are unlikely to full supplant large bilateral and multilateral sources of foreign aid, nor can they serve national foreign policy interests like more standard foreign aid.
+Testimony at a 2017 US Senate Committee on Foreign Relations hearing maintained that "[n]either remittances nor philanthropy can replace the expertise, scale, or agenda setting capacity of American foreign policy assistance" [@Runde:2017].
+Thus, despite promising new research on how to raise funds from private sources for besieged NGOs abroad, official aid donors also need to mount a coherent response to the shrinking civic space.
+
+# References

### 2024-11-02 manuscript.qmd
Add Suparna's stuff

@@
 
+The experiences of organizations such as X are by no means a rare occurrence.
+While a large literature in law and society and political science focuses on the role of international law, domestic legal institutions, and legal mobilization that may reduce repression [@Klug:2005; @PowellStaton:2009; @Simmons:2009; @ConradRitter:2013], the last two decades have made it clear that states are also using law not to expand rights, but to further restrict rights and repress organizations [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @BakkeMitchellSmidt:2020; @GlasiusSchalkDeLange:2020; @BromleySchoferLonghofer:2020; @Chaudhry:2022].
+Administrative crackdown, or the use of laws to create barriers to entry, funding, and advocacy by civil society groups has proliferated across the globe over the past few decades—both across democracies and autocracies [@Chaudhry:2022].
+The latest annual report from CIVICUS Monitor, which assesses the state of civil society worldwide, paints a grim picture.
+As of 2023, a third of the world’s population now lives in countries with closed civic space, and 87% of the world’s population now lives in countries where the space available to civil society organizations (CSOs) is closed, repressed, or obstructed [@CIVICUSMonitor].
+
+The growth in states using law to repress can be attributed to three main factors.
+First, administrative or legal crackdowns have many advantages over the use of violence.
+It allows governments to maintain a democratic facade abroad and limit the domestic backlash because citizens tend to view legal crackdown as regulation rather than repression [@Chaudhry:2022].
+Anti-NGO laws rarely elicit international condemnation or threats to withdraw aid.
+States also learn about the efficacy of these legal strategies from their peers and subsequently implement them in their own country [@GlasiusSchalkDeLange:2020; @Gilbert:2020; @Chaudhry:2022].
+Finally, these tactics are part of a broader trend of using legal means to subvert democratic institutions and norms [@LevitskyZiblatt:2018].
+
+Administrative crackdown via anti-NGO laws that bar access to foreign funding and foreign technical assistance has been particularly insidious.
+Without access to foreign funds, many INGOs cannot continue their overseas operations.
+Even domestic NGOs may be heavily strapped for money.
+Most NGOs in the Global South require a constant stream of competitive grants from foreign donors to be able to implement specific projects [@Bush:2015].
+Local groups tackling contentious issues may not be able to raise funds domestically, as publics may prefer donating to service organizations focusing on health, education, and poverty [@DupuyRonPrakash:2016; @Brechenmacher:2017].
+Even local philanthropists may be deterred from giving to organizations focusing on contentious causes such as advocacy, media freedom, and anti-corruption initiatives due to poor tax incentives or a fear of retribution.
+
+The effects of such repression are visible in many countries.
+For instance, the 2009 Ethiopian Charities and Societies Proclamation Act stipulated that NGOs working on any rights issues in the country must acquire 90% of their funding from domestic sources.
+It also prohibits organizations from spending more than 30% of their budget on "administrative costs," which the act does not define, but could be interpreted to include the provision of free legal aid, advocacy, and other activities essential to the missions of rights groups [@Brechenmacher:2017, 69].
+This law dealt a devastating blow to Ethiopian civil society, as most domestic NGOs working on human rights relied almost exclusively on foreign aid [@Brechenmacher:2017].
+Within just two years, the total number of local NGOs reduced by 25%, and by 90% among human rights NGOs [@DupuyRonPrakash:2016].
+INGOs such as Mercy Corps and Action for Development abandoned their conflict resolution work, while Human Rights Watch, Amnesty International, and the International Federation for Human Rights were prevented from opening field offices in the country.
+Only 10% of the 125 human rights groups in operation at the time of the law’s passage registered under the new requirement [@DupuyRonPrakash:2016].
+
 Given the exponential increase in administrative and legal repression of CSOs across the globe and its effectiveness at muzzling groups, how have donors responded?
@@
 
-## Foreign aid donors
+# Foreign aid donors
 
@@
 Donor agencies rhetorically declare normative justifications for engaging in development, with recipient-focused development goals like poverty reduction, democratization, human rights promotion, and environmental protection [@HoefflerOutram:2011; @Kallman:2020].
-In practice, though, aid is given for far more instrumental reasons based on donor countries' domestic and foreign policy–oriented strategic goals [@AlesinaDollar:2000; @Apodaca:2017; @Schuller:2012], including being responsive to popular democratic pressure [@HeinrichKobayashiLong:2018], insulating themselves from regional instability [@Bermeo:2017], expanding export markets [@BarthelNeumayerNunnenkamp:2014; @Mitchell:2002], influencing recipient states' domestic policies [@Apodaca:2013], and projecting donor country values [@MorgensternBrown:2022].
+In practice, though, aid is given for far more instrumental reasons based on donor countries' domestic and foreign policy-oriented strategic goals [@AlesinaDollar:2000; @Apodaca:2017; @Schuller:2012], including being responsive to popular democratic pressure [@HeinrichKobayashiLong:2018], insulating themselves from regional instability [@Bermeo:2017], expanding export markets [@BarthelNeumayerNunnenkamp:2014], influencing recipient states' domestic policies [@Apodaca:2013], and projecting donor country values [@MorgensternBrown:2022].
 
-### Donor responses and conditionality
+## Donor responses and conditionality
 
@@
 
-### Donor responses to civil society laws
+## State and foundation responses
 
@@
 Donor preferences have also changed.
-Over the past two decades, bilateral donors have increasingly provided short-term grants for projects that have easily measurable, quantifiable outputs [@Bush:2015].
+Over the past two decades, bilateral donors and foundations have increasingly provided short-term grants for projects that have easily measurable, quantifiable outputs [@Bush:2015].
 This model may work well for civic groups working on relatively tame causes or exclusively development-focused goals, but more contentious organizations facing government crackdown may need flexible funding that does not bring immediate results or does not have easily measurable goals [@Herrold:2020].
@@
 
+# Local solutions
+
+Due to the reduced access to (or prohibition on) foreign funds, many local NGOs are also exploring alternative funding models.
+Some NGOs have adapted by engaging in diversification: they create additional affiliate organizations to ensure their survival [@ToeplerPapeBenevolenski:2020].
+They may switch some of their work to another form of organization—including for-profits, which are not subject to the same legal restrictions.
+Creating commercial organizations affiliated with an NGO also allows financial assets to be transferred to an organization that is not subject to the same anti-NGO restrictions.
+In Russia, many formalized environmental NGOs have employed informal models of organization and deregistered with the state due to political and financial constraints.
+They have also diversified and are now working with businesses to accomplish their goals.
+For instance, the Ecological Union in St.
+Petersburg offers private green certification to businesses that undergo a third-party audit of their manufacturing and retail processes [@SundstromHenrySperling:2022].
+
+A growing number of organizations in the Global South are also investing in community philanthropy, women's funds, and developing social enterprises that rely on business revenue.
+Community foundations facilitate democratic and transparent decision making by the local community, which decides who gets grants, and how much.
+This approach allows local groups to respond to (and be held accountable to) local priorities, which also helps overcome many criticisms of international donor aid [@Murad:2014].
+
+Developing social enterprises is another alternative funding model for NGOs.
+Such enterprises are privately owned, for-profit, nonprofit, or hybrid organizations with social objectives.
+They often advance their cause using a business model and then reinvest profits in the community or the nonprofit.
+For instance, Bangladesh has some of the most notable social enterprise nonprofit organizations, including the Bangladesh Rural Advancement Committee (BRAC) and the Grameen Bank.
+BRAC's programs address a wide variety of issues including health, education, craft markets, milk and cattle farming, and micro-credit.
+During the organization's initial stage of development, it reinvested half of its commercial pro ts back into the enterprise and the other half into its nonprofit programming [@ChoSultana:2015].
+Such alternative funding models can thus generate sustainable income while achieving non-profit goals of socioeconomic and community development.
+
+# Continued challenges from anti-NGO rhetoric
+
+Foreign donors already face many criticisms from grantees, including implementing models without taking local contexts into account and prioritizing projects that are more easily quantifiable and regime compatible.
+In an environment of closing civic space, despite many of the adaptations discussed above, NGOs—especially those receiving grants from foreign donors—face even bigger risks.
+In a 2011 survey of 1,473 NGO respondents that received funds from 14 foreign donor agencies, 37% felt the greatest risk of receiving democracy assistance was being labeled an "agent" or "stooge" of outside forces; 48% of respondents lived in countries categorized as "Not Free" by Freedom House [@Barkan:2012].
+In a 2016 survey of over 1,000 activists from Colombia, Egypt, India, Kenya, Russia, Sudan, Turkey, Uganda, Ukraine, and Venezuela, many said that foreign actors often contributed to their security or safety concerns by focusing on digital security to the exclusion of more pressing physical security needs and by failing to take measures to protect partners, including when they have been targeted or arrested [@Miller-Dawkins:2017].
+
+Such risk does not necessarily just imply threats to physical or legal safety.
+Many governments around the world also utilize negative rhetoric that aims to sow distrust between these groups and the communities they serve.
+In Hungary, government officials declared in regards to NGOs that they wanted to "sweep them out of the country," while asking the intelligence services to investigate all NGOs receiving money from the Hungarian-born financier George Soros [@Reuters:2017].
+In rhetoric similar to that in Hungary, where government billboards warn "Don't let George Soros have the last laugh," the Trump administration accused protestors opposing Brett Kavanaugh’s confirmation to the Supreme Court of being funded by wealthy liberal donors like Soros [@Choi:2018].
+
+Recent research shows that NGOs do consider government rhetoric when making operational decisions.
+In a survey experiment of NGOs across Cambodia, Uganda, and Serbia, researchers found that negative NGO rhetoric significantly reduces NGO leaders’ willingness to work in a community.
+It also rescues NGOs’ preferences for partnering with local government authorities [@HollerbauerNagawaRobertson:2024].
+Thus, NGOs prefer to avoid working in communities not just where government authorities repress these groups, but also in areas where authorities use anti-NGO rhetoric.
+While donor responses have largely focused on navigating and adapting to repression, the use of such rhetoric raises broader questions about research and policy on the nonprofit sector: how is such rhetoric changing public attitudes and engagement with these groups?
+Given that the looming threat of repression, and even anti-NGO rhetoric can exacerbate burnout rates [@JoscelyneKnuckeySatterthwaite:2015], what are organizations doing to support human rights advocates, and what kinds of interventions can help support human rights activists and professionals?
+
+In response to these concerns, organizations such as Defend Defenders, Frontline Defenders, Peace Brigades International, Civil Rights Defenders, and Protection International provide physical security, legal and digital support, and mental health resources to activists at risk.
+Regional initiatives such as the Mesoamerican Women Human Rights Defenders Initiative, Forum Asia, and the Euro-Mediterranean Foundation of Support for Human Rights Defenders also help provide formal protection mechanisms for activists.
+
+In many cases, foreign donors have also allied with local NGOs, and provided training to local groups to help defeat anti-NGO bills.
+In Kyrgyzstan, when legislators introduced a draft bill outlining NGO restrictions, Freedom House collaborated with local NGOs to lobby the government.
+They warned legislators that the proposed Russian-style "Foreign Agents" law would adversely affect charitable and humanitarian NGOs and that these restrictions would "lead to a reduction in the number of social services that the population of the country desperately need" [@CheesemanDodsworth:2023, 624].
+At the international level, the US, EU, and DAC-EU members—which provide a quarter to a third of official development assistance to Kyrgyzstan—also made their concerns known to legislators [@CheesemanDodsworth:2023].
+The bill was eventually defeated in 2016.
+Kyrgyz legislators who voted against the bill cited the need for foreign funding: "We get financial assistance from [international organizations] in many fields including healthcare, education, and agriculture among others.
+We need this money" [@Lelik:2016].
+
+Local NGOs similarly partnered with INGOs in Kenya; their campaign emphasized that 8,500 organizations in the country rely heavily on foreign donors and that the proposed 15% cap on foreign funding would have devastating consequences for the NGO sector.
+During the subsequent reading of amendments to the NGO bill, several legislators explained their rejection of the bill in terms of the risks for the socioeconomic development of Kenya and their constituents [@Berger-KernHetzWagner:2021].
+These examples highlight how donors can use their leverage, especially their role in social service provision, to successfully pushback against repressive NGO laws.
+
+# Conclusion
+
+Donors, especially international ones, have long faced criticisms by practitioners, policymakers and academics alike, arguing that they replicate global hierarchies and are unresponsive to the needs of local communities.
+But increasing civic space crackdown, especially in many established democracies in the West, means that when such countries, or donors in these countries speak out or take actions against civil society restrictions abroad, they may have less credibility.
+For instance, the U.S.
+government has teetered dangerously close to the rhetoric and practices shown by many other governments repressing civil society.
+In June 2018, Nikki Haley, U.S.
+Ambassador to the United Nations (UN), blamed U.S.
+withdrawal from the UN Human Rights Council on NGOs, a move that Sarah Morgan, Washington Director of Human Rights Watch, called "astounding." This is intrinsic to the reasons why so far, international actors have struggled to mount a coherent campaign.
+
+While examples of Kenya and Kyrgyzstan discussed above show that donors and local activists can successfully push back against anti-NGO laws, a more systematic approach is needed.
+To tackle this phenomenon more systematically, Western states need a more effective coordinated response that engages not just aid agencies, donors, and international NGOs, but also engages businesses, development communities, diplomatic actors and local populations to deal with larger structural issues that exacerbate government crackdown on civil society.
+Without such an effort, repression of civil society organizations may be the canary in the coal mine that sets the groundwork for future democratic erosion.
+
 # References

### 2024-11-02 manuscript.qmd
MEGA EDITS

--- 2024-11-02 manuscript.qmd
@@
 bibliography: references.bib
-csl: csl/american-sociological-association.csl
+csl: csl/chicago-author-date.csl
 toc: true
@@
     wordcount-banner: true
+    include-in-header:
+      - text: |
+          <style type="text/css">
+          h2.hide {
+            display: none;
+          }
+          </style>
 params:
   wordcount: |
-    <strong>{{< words-sum body-note-ref >}} total words</strong>: {{< words-body >}} in the body • {{< words-ref >}} in the references • {{< words-note >}} in the notes
+    <strong>{{< words-sum body-ref >}} total words</strong>: {{< words-body >}} in the body • {{< words-ref >}} in the references
 ---
@@
 
-Given the exponential increase in administrative and legal repression of CSOs across the globe and its effectiveness at muzzling groups, how have donors responded?
+::: {.content-visible when-format="wordcount-html"}
+# Introduction {.hide}
+:::
+
+Beginning in 2003, Africa and Middle East Refugee Assistance (AMERA) International provided pro bono legal support for refugees entering Egypt.
+Based in London, this international non-governmental organization (INGO) maintained a small staff of legal experts in Cairo and raised the bulk of its funding from organizations in London, receiving the bulk of its revenue from a private charitable foundation, Comic Relief UK.
+Despite laws restricting INGO activities, AMERA enjoyed a positive relationship with government authorities.
+In the wake of the 2011 Arab Uprisings, though, the Egyptian government began strictly enforcing laws prohibiting foreign funding for civil society, blocking AMERA's access to Comic Relief donations.
+In 2014, AMERA shuttered its Egyptian office and redirected its programming to local projects in the United Kingdom [@Heiss:2019a].
+
+The experiences of NGOs like AMERA are common.
+While a large literature in law and political science focuses on the role of international law, domestic legal institutions, and legal mobilization that may reduce repression [@Klug:2005; @PowellStaton:2009; @Simmons:2009; @ConradRitter:2013], the last two decades have shown that states also use laws not to expand rights, but to further restrict rights and repress organizations [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @BakkeMitchellSmidt:2020; @GlasiusSchalkDeLange:2020; @BromleySchoferLonghofer:2020; @Chaudhry:2022].
+Administrative crackdown—or the use of laws to create barriers to entry, funding, and advocacy by civil society groups—has proliferated across the globe over the past few decades—both across democracies and autocracies [@Chaudhry:2022].
+As of 2023, 87% of the world's population now lives in countries where the space available to civil society is closed, repressed, or obstructed [@CIVICUSMonitor].
+
+The growth in states using law to repress can be attributed to three main factors.
+First, administrative or legal crackdowns have many advantages over the use of violence.
+It allows governments to maintain a democratic facade abroad and limit the domestic backlash because citizens tend to view legal crackdown as regulation rather than repression [@Chaudhry:2022].
+Anti-NGO laws rarely elicit international condemnation or threats to withdraw aid.
+States also learn about the efficacy of these legal strategies from their peers and subsequently implement them in their own country [@GlasiusSchalkDeLange:2020; @Gilbert:2020; @Chaudhry:2022].
+Finally, these tactics are part of a broader trend of using legal means to subvert democratic institutions and norms [@LevitskyZiblatt:2018].
+
+Administrative crackdown via anti-NGO laws that bar access to foreign funding and foreign technical assistance has been particularly insidious.
+Without access to foreign funds, many INGOs cannot continue their overseas operations.
+Even domestic NGOs may be heavily strapped for money.
+Most NGOs in the Global South require a constant stream of competitive grants from foreign donors to be able to implement specific projects [@Bush:2015].
+Local groups tackling contentious issues may not be able to raise funds domestically, as publics may prefer donating to service organizations focusing on health, education, and poverty [@DupuyRonPrakash:2016; @Brechenmacher:2017].
+Even local philanthropists may be deterred from giving to organizations focusing on contentious causes such as advocacy, media freedom, and anti-corruption initiatives due to poor tax incentives or a fear of retribution.
+
+The effects of such repression are visible in many countries.
+For instance, the 2009 Ethiopian Charities and Societies Proclamation Act stipulated that NGOs working on any rights issues in the country must acquire 90% of their funding from domestic sources.
+It also prohibited organizations from spending more than 30% of their budget on "administrative costs," which the act does not define, but could be interpreted to include the provision of free legal aid, advocacy, and other activities essential to the missions of rights groups [@Brechenmacher:2017, 69].
+This law dealt a devastating blow to Ethiopian civil society, as most domestic NGOs working on human rights relied almost exclusively on foreign aid [@Brechenmacher:2017].
+Within two years, the total number of local NGOs decreased by 25%, and by an astounding 90% among human rights NGOs [@DupuyRonPrakash:2016].
+INGOs such as Mercy Corps and Action for Development abandoned their conflict resolution work, while Human Rights Watch, Amnesty International, and the International Federation for Human Rights were prevented from opening field offices in the country.
+Of the 125 human rights NGOs in operation when the law passed, only 10% registered under the new requirement [@DupuyRonPrakash:2016].
+
+Given the increase in administrative and legal repression of civil society, how have their donors responded?
 How do these responses differ across foreign government and private donors?
@@
 
-## Foreign aid donors
+# State responses
 
-A rich literature on the purposes of foreign aid explores a broad collection of reasons why states transfer money and resources to others.
-Aid agencies and international organizations face competing interests when deciding how to allocate resources, which projects to undertake, and which countries to work with.
-Donor agencies rhetorically declare normative justifications for engaging in development, with recipient-focused development goals like poverty reduction, democratization, human rights promotion, and environmental protection [@HoefflerOutram:2011; @Kallman:2020].
-In practice, though, aid is given for far more instrumental reasons based on donor countries' domestic and foreign policy–oriented strategic goals [@AlesinaDollar:2000; @Apodaca:2017; @Schuller:2012], including being responsive to popular democratic pressure [@HeinrichKobayashiLong:2018], insulating themselves from regional instability [@Bermeo:2017], expanding export markets [@BarthelNeumayerNunnenkamp:2014; @Mitchell:2002], influencing recipient states' domestic policies [@Apodaca:2013], and projecting donor country values [@MorgensternBrown:2022].
-
-### Donor responses and conditionality
-
-A key enforcement mechanism to ensure that recipient-country actions align with donor-country goals—and thus reduce the moral hazard inherent in donating resources—is to condition aid on the achievement of specific democratizing and economic goals [@Winters:2024; @CheesemanSwedlundOBrien-Udry:2024; @HoefflerJustino:2024; @Kilby:2024; @Koch:2015].
-Many donor governments condition aid on human rights outcomes, using progress in respect for human rights as leverage to shape recipient-country behavior.
-Conditional aid sends dual signals: it serves as an instrumental "stick" to pressure governments to change policies and an expressive "flag" to signal donor commitment to human rights norms [@DasandiErez:2023].
+A rich literature on the purposes of foreign aid explores why states transfer money and resources to others.
+Bilateral national aid agencies and multilateral international organizations face competing interests when deciding how to allocate resources, which projects to undertake, and which countries to work with.
+States rhetorically declare normative justifications for engaging in development, with recipient-focused development goals like poverty reduction, democratization, human rights promotion, and environmental protection [@HoefflerOutram:2011].
+In practice, though, aid is given for far more instrumental reasons based on donor countries' domestic and foreign policy-oriented strategic goals [@AlesinaDollar:2000; @Apodaca:2017].
+A key enforcement mechanism to ensure that recipient-country actions align with donor-country goals is to condition aid on the achievement of specific democratizing and economic goals [@CheesemanSwedlundOBrien-Udry:2024; @HoefflerJustino:2024].
 However, this dual-purpose strategy often creates dilemmas—if aid is withdrawn, it may jeopardize humanitarian programs or worsen political conditions.
-Conversely, continuing aid might signal tolerance for rights violations, weakening the credibility of the donor’s human rights agenda.
-
-The often-conflicting trends of both pursuing normative development goals and furthering foreign policy interests leads to mixed responses when recipient countries fail to adhere to aid conditions and violate human rights.
-Prior research finds conflicting relationships between human rights abuses and foreign aid, variously concluding that when confronted by recipient states that violate human rights, donor agencies punish repressive states by reducing, suspending, or withdrawing aid [@LebovicVoeten:2009], increasing aid (or withdrawing less aid) to repressive states as a way to counter state violence [@Demirel-PeggMoskowitz:2009; @Nielsen:2013; @HeinrichKobayashiLong:2018; @EsareyDeMeritt:2017], or make no measurable change in aid allocations [@Neumayer:2003; @Neumayer:2003b].
+Conversely, continuing aid might signal tolerance for rights violations, weakening the credibility of the donor's human rights agenda.
+Prior research finds conflicting relationships between human rights abuses and foreign aid, variously concluding that donor agencies punish repressive states by reducing or withdrawing aid [@LebovicVoeten:2009], increasing aid to repressive states as a way to counter state violence [@Nielsen:2013], or making no measurable change in aid allocations [@Neumayer:2003; @Neumayer:2003b].
 
@@
 Donor countries have a range of possible responses to recipient repression, including doing nothing and ignoring the violation, threatening to change aid allocations, withdrawing or suspending aid, or shifting aid to other projects or programs.
-In turn, recipient governments can then respond by doing nothing and ignoring the threat or suspension, change their behavior and cease the repression to improve their relationship with the donor country, or pressure the donor country to back down and withdraw their complaint.
-In this simplified decision tree, there are 12 possible combinations of outcomes,^[@CheesemanSwedlundOBrien-Udry:2024 describe additional stages in their tree, with donors responding to recipients' responses, creating dozens of possible pathways.] and each pathway is conditional on a host of donor-country and recipient-country characteristics and contexts, including geopolitical dynamics in the recipient country's region, domestic political institutions within the donor country, public opinion in the donor country, coordination between coalitions of donors, the type of aid, and the type of human rights violation, among others [@CheesemanSwedlundOBrien-Udry:2024].
-Such complexity makes it difficult to find consistent treatment effects in research that explores the overall effect of repression on aid—that is, we cannot definitively say that human rights causes an overall reduction or increase in aid.
+In turn, recipient governments can then respond by doing nothing and ignoring the threat or suspension, changing their behavior and ceasing the repression to improve their relationship with the donor country, or pressuring the donor country to back down and withdraw their complaint.
+In this simplified decision tree, there are 12 possible combinations of outcomes, and each pathway is conditional on a host of donor-country and recipient-country characteristics and contexts, including geopolitical dynamics in the recipient country's region, domestic political institutions within the donor country, public opinion in the donor country, coordination between coalitions of donors, the type of aid, and the type of human rights violation, among others [@CheesemanSwedlundOBrien-Udry:2024].
+Such complexity makes it difficult to find consistent treatment effects in research that explores the overall effect of repression on aid—that is, we cannot definitively say that human rights abuses cause an overall reduction or increase in aid.
 Instead, newer research looks at donor responses in more specific circumstances, exploring the factors that lead to specific response pathways [@Corwin:2023].
 
-### Donor responses to civil society laws
-
-Research on aid conditionality and agency responses to human rights abuses tends to look at more overt and violent forms of political oppression like respect for personal integrity, civil, and political rights [@Corwin:2023; @Neumayer:2003; @CheesemanSwedlundOBrien-Udry:2024], including arresting political activists, closing media outlets, suppressing protests, engaging in extrajudicial killings, and so on.
-The effects of more subtle administrative crackdown on aid, including the passage of anti-NGO laws and the constricting of civic space, are less explored and understood.
-
-Civil society plays a crucial role in the distribution and implementation of international aid.
-Governmental aid agencies, such as the United States Agency for International Development (USAID) and the United Kingdom's Foreign, Commonwealth and Development Office (FCDO), channel substantial funding through NGOs in recipient countries.
-TODO: Something about the amount of money that goes through CS, perhaps highlighting just the US or UK.
-Larger international organizations also rely on domestic NGOs to carry out humanitarian and development projects.
-TODO: Something about the amount of money that the World Bank or UNHCR sends to NGOs.
+Research on aid conditionality and agency responses to human rights abuses tends to look at more overt and violent forms of political oppression.
+The effects of more subtle administrative crackdown on aid, including the passage of anti-NGO laws and the constricting of civic space, are less explored.
+Civil society plays a central role in the distribution and implementation of international aid.
+Governmental agencies, such as the United States Agency for International Development (USAID) and the United Kingdom's Foreign, Commonwealth and Development Office (FCDO), channel substantial funding through NGOs in recipient countries.
 Outsourcing the provision of aid to NGOs allows donors to avoid aid capture in weak institutional environments and can lead to more effective program implementation [@Dietrich:2013; @CruzLabonneWright:2024].
-Given the importance of NGOs in aid delivery, we have reason to expect that donor responses to civil society crackdown abroad are treated differently than more violent forms of repression, and thus likely to follow different pathways in @fig-agency-response-tree.
+Given the importance of NGOs in aid delivery, we have reason to expect that donor responses to civil society crackdown abroad are treated differently than more violent forms of repression, following different pathways in @fig-agency-response-tree.
 
@@
 @ChristensenWeinstein:2013 show that aid from bilateral donors—or direct country-to-country aid—sees an average decrease of $25–\$50 million after recipient states pass laws that restrict the flow of foreign funds to NGOs.
-However, aid from international, multilateral agencies is not affected by these laws.
-Similarly @DupuyPrakash:2018 find that the adoption of a restrictive foreign funding law is associated with a 32% decline in aid from national, bilateral donors, but has no corresponding change in aid from international, multilateral donors.
+However, aid from multilateral agencies is not affected by these laws.
+Similarly, @DupuyPrakash:2018 find that the adoption of a restrictive foreign funding law is associated with a 32% decline in aid from bilateral donors, but has no corresponding change in aid from multilateral donors.
 
-One key reason for the divergence in the responses of bilateral and multilateral donors is the degree to which these donors rely on recipient-country NGOs to implement programs.
-Donor governments and agencies have different experiences with NGOs and correspondingly prioritize the openness of civil society in recipient countries in different ways.
+One key reason for the divergence in the responses of national and international agencies is the degree to which these donors rely on recipient-country NGOs to implement programs.
+Donor agencies prioritize the openness of civil society in recipient countries in different ways.
 Multilateral donors like the World Bank and OECD tend to channel money directly to recipient governments rather than domestic NGOs and are thus more insulated from the effects of local civil society repression, which likely explains the lack of formal response to anti-NGO laws [@ChristensenWeinstein:2013; @DupuyPrakash:2018].
 There is more variation, however, in how bilateral donors work with domestic NGOs.
-For example, @AllenFlynn:2018 find that the domestic political environments of donor countries shape how those countries channel their aid abroad.
+@AllenFlynn:2018 find that the domestic political environments of donor countries shape how those countries channel their aid abroad.
 Because they tend to focus more on poverty alleviation initiatives, states with left-leaning governments typically channel more aid through recipient-country NGOs.
@@
 The nature of civil society restrictions also influences how official donor agencies respond.
-In their initial work on anti-NGO restrictions, @ChristensenWeinstein:2013 cataloged civil society laws into three broader categories—barriers to funding, barriers to entry, and barriers to advocacy—but they, and others like @DupuyPrakash:2018, dedicated the bulk of their analysis to looking at the effects of barriers to funding.
+@ChristensenWeinstein:2013 cataloged civil society laws into three broader categories—barriers to funding, barriers to entry, and barriers to advocacy—but they, and others like @DupuyPrakash:2018, dedicated the bulk of their analysis to looking at the effects of barriers to funding.
 This focus on funding restrictions is warranted, as most authoritarian civil society legal restrictions have been aimed specifically at reducing or controlling foreign funding [@Heiss:2019; @Carothers:2015a; @CarothersBrechenmacher:2014].
-Moreover, the relationship between foreign aid and foreign funding restrictions appears to be bidirectional, as states are more likely to pass restrictive foreign funding laws in response to increases in bilateral aid [@DupuyRonPrakash:2016], especially if that aid arrives during competitive elections or times of political protest or instability [@Chaudhry:2022; @Heiss:2017].
+Moreover, the relationship between foreign aid and foreign funding restrictions appears to be bidirectional, as states are more likely to pass restrictive foreign funding laws in response to increases in bilateral aid [@DupuyRonPrakash:2016], especially during competitive elections or times of political protest or instability [@Chaudhry:2022; @Heiss:2017].
 Foreign funding restrictions and aid are thus linked: increasing aid leads to more funding restrictions [@DupuyRonPrakash:2016], which in turn lead to reductions in aid [@DupuyPrakash:2018].
 
-Though the majority of anti-NGO laws have targeted foreign funding, other laws seek to impose barriers to entry, such as strict and burdensome registration requirements, and barriers to advocacy that explicitly limit NGO political participation [@Chaudhry:2022].
+Though most anti-NGO laws have targeted foreign funding, others impose barriers to entry, such as strict and burdensome registration requirements, and barriers to advocacy that explicitly limit NGO political participation [@Chaudhry:2022].
 Donor countries respond differently to these types of oppression.
@@
 
-Overall bilateral aid does not decrease significantly in response to new barriers to entry, but the distribution channels of that aid shift as donor countries channel more aid through domestic rather than foreign NGOs [@ChaudhryHeiss:2024].
+Overall, bilateral aid does not decrease significantly in response to new barriers to entry, but the distribution channels of that aid shift as donor countries channel more aid through domestic rather than foreign NGOs [@ChaudhryHeiss:2024].
 USAID even explicitly encourages its program managers to rely more on domestic, in-country NGOs when facing a restrictive legal environment [@USAID:2014].
-In addition to changing the amounts and recipients of foreign aid, donor agencies often adjust the focus of the programs they fund, redirecting funding away from programs related to politically sensitive issues like human rights, anti-corruption, and democratic development and towards tamer topics that are more compatible with recipient-country preferences, such as education, health, and humanitarian services [@Bush:2015; @CheesemanDesrosiers:2023].
+In addition to changing the amounts and recipients of foreign aid, donor agencies often adjust the focus of the programs they fund, redirecting funding away from programs related to politically sensitive issues like human rights, anti-corruption, and democratic development and towards tamer topics that are more compatible with recipient-country preferences, such as education, health, and humanitarian services [@Bush:2015].
 
 These shifts in funding patterns in response to restrictive civil society laws have had measurable effects on NGOs in recipient countries.
-Samantha Power, current administrator of USAID, has pledged to send 25% of USAID funding to local NGOs over the next few years [@Power:2021], though practical obstacles remain.
+In 2021, USAID pledged to send 25% of its aid through local NGOs over the next few years [@Power:2021], but practical obstacles remain.
 New local partners take longer to vet and train compared to established INGO partners.
-Turning to local organizations entails greater risk, as domestic NGOs can be easily shut down based on the whims of the government.
+Turning to local organizations entails greater risk, as domestic NGOs can be easily shut down by their governments.
 Additionally, communities often distrust local groups that accept Western aid, especially those focused on democracy assistance [@Barkan:2012].
 Donor preferences have also changed.
-Over the past two decades, bilateral donors have increasingly provided short-term grants for projects that have easily measurable, quantifiable outputs [@Bush:2015].
+Over the past two decades, bilateral donors and foundations have increasingly provided short-term grants for projects that have easily measurable, quantifiable outputs [@Bush:2015].
 This model may work well for civic groups working on relatively tame causes or exclusively development-focused goals, but more contentious organizations facing government crackdown may need flexible funding that does not bring immediate results or does not have easily measurable goals [@Herrold:2020].
 
-## Private donors
+# Private donor responses
 
-As explored above, given the centrality of aid in domestic foreign policy decisions and economic development strategy, substantial research in development economics, political science, and international relations has looked at the motivations behind bilateral and multilateral foreign aid, especially in response to human rights abuses in recipient states.
-However, states and international organizations are not the only institutions that engage in international giving.
+States and international organizations are not the only institutions that engage in international giving.
 Private individuals exercise power and influence in unique ways that differ from other non-state actors.
 Since 2000, foreign aid funding from philanthropic foundations and individual donors in the United States has more than doubled [@DesaiKharas:2018], with nearly \$30 billion donated to international affairs and development organizations in 2023 [@GivingUSA:2024].
-This growth in private foreign aid has had a substantial effect on overall international  aid flows and often serves as a complement to official bilateral and multilateral aid, particularly in response to acute humanitarian crises [@DesaiKharas:2018].
-
-Understanding the dynamics of supporting international causes is important as philanthropy towards such causes can be harder to motivate.
-Work in nonprofit and philanthropic studies has long explored the personal motivations of individual donors, including reputation, altruism, and psychological benefits [see @BekkersWiepking:2011 for an exploration of eight different motivating mechanisms], but less research—particularly in political science and international  relations—has been done regarding the organizational or institutional mechanisms that drive donor behavior.
+Private foreign aid now serves as a complement to official bilateral and multilateral aid, particularly in response to acute humanitarian crises [@DesaiKharas:2018].
+Work in nonprofit and philanthropic studies has long explored the personal motivations of individual donors, including reputation, altruism, and psychological benefits [@BekkersWiepking:2011], but there is far less research regarding the organizational or institutional mechanisms that drive donor behavior.
 Donors are motivated by the broader political and institutional contexts of NGO programming, and individuals are less likely to donate to foreign nonprofits and prefer giving to local causes, since the result of their giving is more visible [@CasaleBaumann:2015; @Tremblay-BoirePrakash:2017; @Wiepking:2010].
@@
 Donor motivations influence how individuals respond when potential recipient organizations face legal trouble abroad.
-However, the exact menu of responses is more limited than what is available to bilateral and multilateral states, as seen previously in @fig-individual-response-tree, and far more understudied.
+However, the exact menu of responses is more limited (and less explored) than what is available to bilateral and multilateral states.
 @fig-individual-response-tree presents a preliminary framework for outlining possible individual donor responses to anti-NGO repression.
-When a potential recipient NGO faces legal crackdown in its host country, thus limiting its operations, interested individual donors can respond by (1) increasing their donations, as a sign of support and solidarity, (2) decreasing their donations, punishing the NGO for doing something to run afoul of its host government, or (3) not considering the host country legal environment and making no change in their donation behavior.
+When a potential recipient NGO faces legal crackdown in its host country, private donors can respond by (1) increasing their donations as a sign of support and solidarity, (2) decreasing their donations, punishing the NGO for doing something to run afoul of its host government, or (3) not considering the host country legal environment and making no change in their donation behavior.
 @ChaudhryDotsonHeiss:2021 explore how different combinations of donor characteristics determine which response individuals are likely to take when NGOs face crackdown.
@@
 
-Additional experimental evidence examines how donors react to the knowledge that potential recipient NGOs face legal difficulties.
-In a vignette-based factorial experiment, @ChaudhryHeiss:2021 find that foreign civil society restrictions serve as a heuristic when deciding to donate—learning about NGO repression increases individual generosity and causes potential donors to be willing to give substantially more to legally restricted nonprofits.
+Experimental evidence examines how donors react to the knowledge that potential recipient NGOs face legal difficulties.
+@ChaudhryHeiss:2021 find that foreign civil society restrictions serve as a heuristic when deciding to donate—learning about NGO repression increases individual generosity and causes potential donors to be willing to give substantially more to legally restricted nonprofits.
 This effect is strongest for privately-funded human rights-focused NGOs that face legal crackdown, as individuals feel that the organization deserves their support.
-Several participants justified their increased support because the experimental organization is "doing good work in countries where it is tough for groups like them to operate and they
-need all the help they can get" [@ChaudhryHeiss:2021, 496].
-Building on this work, @ChaudhryDotsonHeiss:2024 use a survey experiment to examine how organizational characteristics (e.g., issue area, funding sources) and practices (e.g., financial transparency) interact with structural concerns like legal crackdown abroad.
+Several participants justified their increased support because restricted NGOs are "doing good work in countries where it is tough for groups like them to operate and they need all the help they can get" [@ChaudhryHeiss:2021, 496].
+Building on this work, @ChaudhryDotsonHeiss:2024 examine how organizational characteristics (e.g., issue area, funding sources) and practices (e.g., financial transparency) interact with structural concerns like legal crackdown abroad.
 They find that, all else equal, legal crackdown on its own causes donors to reduce their support for nonprofits—individuals are 5–10 percentage points less likely to donate to an NGO that is criticized or restricted, compared to an organization that enjoys a friendly relationship with its host government.
@@
 Financial transparency and accountability protect each increase the probability of donations by nine percentage points under the worst conditions of legal crackdown.
-Individual donors thus typically reduce their support when seeing legal crackdown, but change their response and increase their when seeing other signs that the organization follows best practices and is more deserving.
+Individual donors thus typically reduce their support when seeing legal crackdown, but change their response and increase their support when seeing other signs that the organization follows best practices and is more deserving.
 
 The question of how organizations respond to different donor reactions to legal crackdown remains unexplored (see the far right panel of @fig-individual-response-tree).
-To our knowledge, no work has been done regarding individual-level nonprofit fundraising strategies in the face of closing civic space, opening promising avenues for future research.
+To our knowledge, no work has been done regarding individual-level nonprofit fundraising strategies in the face of closing civic space, presenting promising avenues for future research.
 
-While private donations have become an important source of funding in the age of closing civi space and increased restrictions on official foreign funding, they are unlikely to full supplant large bilateral and multilateral sources of foreign aid, nor can they serve national foreign policy interests like more standard foreign aid.
-Testimony at a 2017 US Senate Committee on Foreign Relations hearing maintained that "[n]either remittances nor philanthropy can replace the expertise, scale, or agenda setting capacity of American foreign policy assistance" [@Runde:2017].
+While private donations have become an important source of funding in the age of closing civic space and increased restrictions on official foreign funding, they are unlikely to full supplant large bilateral and multilateral sources of foreign aid, nor can they serve national foreign policy interests like more standard foreign aid.
+Testimony at a 2017 U.S.
+Senate Committee on Foreign Relations hearing maintained that "[n]either remittances nor philanthropy can replace the expertise, scale, or agenda setting capacity of American foreign policy assistance" [@Runde:2017].
 Thus, despite promising new research on how to raise funds from private sources for besieged NGOs abroad, official aid donors also need to mount a coherent response to the shrinking civic space.
 
+# Local solutions
+
+Due to the reduced access to (or prohibition on) foreign funds, many domestic NGOs have explored alternative funding models.
+Some have engaged in diversification, creating affiliate organizations to ensure their survival [@ToeplerPapeBenevolenski:2020].
+Others have switched organizational forms, becoming for-profit firms that are not subject to the same legal restrictions [@YeHeiss:2024].
+In Russia, many environmental NGOs have employed informal models of organization and deregistered with the state due to political and financial constraints, while others partner with businesses directly—for instance, the Ecological Union in St.
+Petersburg offers private green certification to businesses that undergo a third-party audit of their manufacturing and retail processes [@SundstromHenrySperling:2022].
+
+A growing number of organizations in the Global South have turned to community philanthropy.
+Community foundations facilitate democratic and transparent decision-making by allowing the local community to decide who recieves grants.
+This approach allows local groups to respond and be held accountable to local priorities, overcoming many criticisms of international donor aid [@Murad:2014].
+
+Social enterprises—for-profit, nonprofit, or hybrid organizations with social objectives—offer another alternative funding model for NGOs.
+These entities often advance their cause with a for-profit business model and reinvest profits in the community or the nonprofit.
+For instance, Bangladesh has some of the most notable social enterprise nonprofit organizations, including the Bangladesh Rural Advancement Committee (BRAC) and the Grameen Bank.
+BRAC's programs address a wide variety of issues, including health, education, craft markets, milk and cattle farming, and micro-credit.
+During the organization's initial stage of development, it reinvested half of its commercial profits back into the enterprise and the other half into its nonprofit programming [@ChoSultana:2015].
+These alternative funding models can generate sustainable income while achieving social goals of socioeconomic and community development.
+
+# Continued challenges from anti-NGO rhetoric
+
+Foreign donors already face many criticisms from grantees, including implementing models without taking local contexts into account and prioritizing projects that are more easily quantifiable and regime compatible.
+In an environment of closing civic space, NGOs—especially those receiving grants from foreign donors—face even bigger risks.
+In a 2011 survey of 1,473 NGO respondents that had received foreign funds, 37% felt the greatest risk of receiving democracy assistance was being labeled an "agent" or "stooge" of outside forces.
+[@Barkan:2012].
+In a 2016 survey of over 1,000 activists from Colombia, Egypt, India, Kenya, Russia, Sudan, Turkey, Uganda, Ukraine, and Venezuela, respondents reported that their foreign donors created security and safety concerns by neglecting to take adequate measures to protect their partners, even when they have been targeted or arrested [@Miller-Dawkins:2017].
+
+Such risk does not necessarily just imply threats to physical or legal safety.
+Many states use negative rhetoric designed to sow distrust between these groups and the communities they serve.
+In Hungary, government officials declared they wanted to "sweep [NGOs] out of the country," while asking the intelligence services to investigate all NGOs receiving money from the Hungarian-born financier George Soros [@Reuters:2017].
+Using similar rhetoric, the Trump administration accused protestors opposing Brett Kavanaugh's confirmation to the Supreme Court of being funded by wealthy liberal donors like Soros [@Choi:2018].
+
+Recent research shows that NGOs consider government rhetoric when making operational decisions.
+A survey experiment of NGOs across Cambodia, Uganda, and Serbia, found that negative NGO rhetoric significantly reduces NGO leaders' willingness to work in a community and curtails NGOs' preferences for partnering with local government authorities [@HollerbauerNagawaRobertson:2024].
+Thus, NGOs not only seek to avoid working in communities repressed by the government, but also avoid working in areas where authorities use anti-NGO rhetoric.
+While donor responses have largely focused on navigating and adapting to repression, the use of such rhetoric raises broader questions about research and policy on the nonprofit sector: how is such rhetoric changing public attitudes and engagement with these groups?
+Given that anti-NGO repression and rhetoric can exacerbate burnout rates [@JoscelyneKnuckeySatterthwaite:2015], what are organizations doing to support human rights advocates, and what kinds of interventions can help support activists and professionals?
+In response to these concerns, organizations such as Defend Defenders, Frontline Defenders, Peace Brigades International, Civil Rights Defenders, and Protection International provide physical security, legal and digital support, and mental health resources to activists at risk.
+Regional initiatives such as the Mesoamerican Women Human Rights Defenders Initiative, Forum Asia, and the Euro-Mediterranean Foundation of Support for Human Rights Defenders also help provide formal protection mechanisms for activists.
+
+In many cases, foreign donors have also allied with local NGOs and provided training to local groups to help defeat anti-NGO bills.
+In Kyrgyzstan, Freedom House collaborated with local NGOs to lobby against a draft bill restricting NGOs.
+They warned legislators that the proposed Russian-style "Foreign Agents" law would harm charitable and humanitarian NGOs and reduce essential social services.
+The US, EU, and DAC-EU members, which provide a quarter to a third of Kyrgyzstan's aid, also expressed concerns directly to Kyrgyz legislators [@CheesemanDodsworth:2023].
+The bill was defeated in 2016, and legislators who voted against the bill cited the country's need for foreign funding: "We get financial assistance from [international organizations] in many fields including healthcare, education, and agriculture among others.
+We need this money" [@Lelik:2016].
+Local NGOs similarly partnered with INGOs in Kenya—their campaign emphasized that 8,500 organizations in the country rely heavily on foreign donors and that a proposed 15% cap on foreign funding would have devastating consequences for the NGO sector.
+During the subsequent reading of amendments to the NGO bill, several legislators justified their rejection of the bill based on the risks posed to the socioeconomic development of Kenya.
+[@Berger-KernHetzWagner:2021].
+These examples highlight how donors can use their leverage, especially their role in social service provision, to successfully push back against repressive NGO laws.
+
+# Conclusion
+
+Donors have long faced criticisms from practitioners, policymakers, and academics, who argue that international donors in particular replicate global hierarchies and are unresponsive to the needs of local communities.
+However, with the global crackdown on civic space—including in many established democracies in the Global North—donors from these countries may lose credibility when speaking out against civil society restrictions abroad.
+For instance, the U.S.
+government has teetered dangerously close to the rhetoric and practices of many other governments repressing civil society.
+In June 2018, Nikki Haley, the U.S.
+Ambassador to the United Nations (UN), blamed NGOs for the U.S.
+withdrawal from the UN Human Rights Council.
+International actors have struggled to mount a coherent campaign against civil society repression.
+
+While examples of Kenya and Kyrgyzstan discussed above show that donors and local activists can successfully push back against anti-NGO laws, a more systematic approach is needed.
+To tackle this administrative repression, Western states need a more effective and coordinated response that engages not only aid agencies, donors, and international NGOs but also businesses, development communities, diplomatic actors, and local populations to address larger structural issues that exacerbate government crackdowns on civil society.
+Without such an effort, the repression of civil society organizations may be the canary in the coal mine that sets the groundwork for future democratic erosion.
+
 # References

### 2024-11-05 manuscript.qmd
Final Suparna edits

@@
 Beginning in 2003, Africa and Middle East Refugee Assistance (AMERA) International provided pro bono legal support for refugees entering Egypt.
-Based in London, this international non-governmental organization (INGO) maintained a small staff of legal experts in Cairo and raised the bulk of its funding from organizations in London, receiving the bulk of its revenue from a private charitable foundation, Comic Relief UK.
-Despite laws restricting INGO activities, AMERA enjoyed a positive relationship with government authorities.
+Based in London, this international non-governmental organization (INGO) maintained a small staff of legal experts in Cairo and raised the bulk of its funding from organizations in London, receiving most of its revenue from a private charitable foundation, Comic Relief UK.
+Despite Egyptian laws restricting INGO activities, AMERA enjoyed a positive relationship with government authorities.
 In the wake of the 2011 Arab Uprisings, though, the Egyptian government began strictly enforcing laws prohibiting foreign funding for civil society, blocking AMERA's access to Comic Relief donations.
@@
 
-The experiences of NGOs like AMERA are common.
-While a large literature in law and political science focuses on the role of international law, domestic legal institutions, and legal mobilization that may reduce repression [@Klug:2005; @PowellStaton:2009; @Simmons:2009; @ConradRitter:2013], the last two decades have shown that states also use laws not to expand rights, but to further restrict rights and repress organizations [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @BakkeMitchellSmidt:2020; @GlasiusSchalkDeLange:2020; @BromleySchoferLonghofer:2020; @Chaudhry:2022].
+The experiences of non-governmental organizations (NGOs) like AMERA are not uncommon.
+While a large literature in law and political science focuses on the role of international law, domestic legal institutions, and legal mobilization that may reduce state crackdown [@Klug:2005; @PowellStaton:2009; @Simmons:2009; @ConradRitter:2013], the last two decades have shown that states also use laws not to expand rights, but to further restrict rights and repress organizations [@ChristensenWeinstein:2013; @DupuyRonPrakash:2016; @BakkeMitchellSmidt:2020; @GlasiusSchalkDeLange:2020; @BromleySchoferLonghofer:2020; @Chaudhry:2022].
 Administrative crackdown—or the use of laws to create barriers to entry, funding, and advocacy by civil society groups—has proliferated across the globe over the past few decades—both across democracies and autocracies [@Chaudhry:2022].
@@
 First, administrative or legal crackdowns have many advantages over the use of violence.
-It allows governments to maintain a democratic facade abroad and limit the domestic backlash because citizens tend to view legal crackdown as regulation rather than repression [@Chaudhry:2022].
+It allows governments to maintain a democratic facade and limit the domestic backlash because citizens tend to view legal crackdown as regulation rather than repression [@Chaudhry:2022].
 Anti-NGO laws rarely elicit international condemnation or threats to withdraw aid.
@@
 Administrative crackdown via anti-NGO laws that bar access to foreign funding and foreign technical assistance has been particularly insidious.
-Without access to foreign funds, many INGOs cannot continue their overseas operations.
+Without access to foreign funds, many INGOs, such as AMERA, cannot continue their overseas operations.
 Even domestic NGOs may be heavily strapped for money.
-Most NGOs in the Global South require a constant stream of competitive grants from foreign donors to be able to implement specific projects [@Bush:2015].
+Most NGOs in the Global South require a constant stream of competitive grants from foreign donors to implement specific projects [@Bush:2015].
 Local groups tackling contentious issues may not be able to raise funds domestically, as publics may prefer donating to service organizations focusing on health, education, and poverty [@DupuyRonPrakash:2016; @Brechenmacher:2017].
-Even local philanthropists may be deterred from giving to organizations focusing on contentious causes such as advocacy, media freedom, and anti-corruption initiatives due to poor tax incentives or a fear of retribution.
+Even local philanthropists may be deterred from giving to organizations focusing on contentious issues such as advocacy, media freedom, and anti-corruption initiatives due to poor tax incentives or a fear of retribution.
 
@@
 
-# State responses
+# Responses from official aid donors
 
-A rich literature on the purposes of foreign aid explores why states transfer money and resources to others.
-Bilateral national aid agencies and multilateral international organizations face competing interests when deciding how to allocate resources, which projects to undertake, and which countries to work with.
-States rhetorically declare normative justifications for engaging in development, with recipient-focused development goals like poverty reduction, democratization, human rights promotion, and environmental protection [@HoefflerOutram:2011].
+A rich literature on the purposes of foreign aid explores why states transfer money and resources to other countries.
+Both national aid agencies and multilateral international organizations face competing interests when deciding how to allocate resources, which projects to undertake, and which countries to work with.
+States rhetorically declare normative justifications for engaging in development, with recipient-focused goals like poverty reduction, democratization, human rights promotion, and environmental protection [@HoefflerOutram:2011].
 In practice, though, aid is given for far more instrumental reasons based on donor countries' domestic and foreign policy-oriented strategic goals [@AlesinaDollar:2000; @Apodaca:2017].
@@
 
-One reason for such mixed results is that the decision to withdraw or suspend aid in response to human rights abuses is often seen as a binary choice driven by self-interested motivations: a donor country that sees human rights abuses in a recipient state will consider its own foreign policy interests before deciding to either continue or curtail aid.
+One reason for such mixed results is that the decision to withdraw or suspend aid in response to human rights abuses (including civil society crackdown) is often seen as a binary choice driven by self-interested motivations: a donor country that sees human rights abuses in a recipient state will consider its own foreign policy interests before deciding to either continue or curtail aid.
 In reality, however, the decision is far more complex and involves a wide range of possible motivations and responses.
@@
 
-Donor countries have a range of possible responses to recipient repression, including doing nothing and ignoring the violation, threatening to change aid allocations, withdrawing or suspending aid, or shifting aid to other projects or programs.
-In turn, recipient governments can then respond by doing nothing and ignoring the threat or suspension, changing their behavior and ceasing the repression to improve their relationship with the donor country, or pressuring the donor country to back down and withdraw their complaint.
+Donor countries thus have a range of possible responses to recipient repression, including doing nothing and ignoring the violation, threatening to change aid allocations, withdrawing or suspending aid, or shifting aid to other projects or programs.
+In turn, recipient governments can then respond by doing nothing and ignoring the suspension, or changing their behavior and ceasing the repression to improve their relationship with the donor country, or pressuring the donor country to back down and withdraw their complaint.
 In this simplified decision tree, there are 12 possible combinations of outcomes, and each pathway is conditional on a host of donor-country and recipient-country characteristics and contexts, including geopolitical dynamics in the recipient country's region, domestic political institutions within the donor country, public opinion in the donor country, coordination between coalitions of donors, the type of aid, and the type of human rights violation, among others [@CheesemanSwedlundOBrien-Udry:2024].
@@
 
-Research on aid conditionality and agency responses to human rights abuses tends to look at more overt and violent forms of political oppression.
+Research on aid conditionality and agency responses to human rights abuses tends to look at more overt and violent forms of political repression.
 The effects of more subtle administrative crackdown on aid, including the passage of anti-NGO laws and the constricting of civic space, are less explored.
 Civil society plays a central role in the distribution and implementation of international aid.
-Governmental agencies, such as the United States Agency for International Development (USAID) and the United Kingdom's Foreign, Commonwealth and Development Office (FCDO), channel substantial funding through NGOs in recipient countries.
+Government agencies, such as the United States Agency for International Development (USAID) and the United Kingdom's Foreign, Commonwealth and Development Office (FCDO), channel substantial funding through NGOs in recipient countries.
 Outsourcing the provision of aid to NGOs allows donors to avoid aid capture in weak institutional environments and can lead to more effective program implementation [@Dietrich:2013; @CruzLabonneWright:2024].
@@
 
-Though most anti-NGO laws have targeted foreign funding, others impose barriers to entry, such as strict and burdensome registration requirements, and barriers to advocacy that explicitly limit NGO political participation [@Chaudhry:2022].
+Though most anti-NGO laws have targeted foreign funding, others impose barriers to entry, such as strict and burdensome registration requirements, and barriers to advocacy that explicitly limit NGO political participation, specifically through vaguely defined "political activities" [@Chaudhry:2022].
 Donor countries respond differently to these types of oppression.
@@
 
-# Private donor responses
+# Responses from private donors
 
@@
 
-While private donations have become an important source of funding in the age of closing civic space and increased restrictions on official foreign funding, they are unlikely to full supplant large bilateral and multilateral sources of foreign aid, nor can they serve national foreign policy interests like more standard foreign aid.
+While private donations have become an important source of funding in the age of closing civic space and increased restrictions on foreign funding, they are unlikely to fully supplant large bilateral and multilateral sources of foreign aid, nor can they serve national foreign policy interests like more standard foreign aid.
 Testimony at a 2017 U.S.
@@
 
-A growing number of organizations in the Global South have turned to community philanthropy.
+A growing number of organizations in the Global South have also turned to community philanthropy.
 Community foundations facilitate democratic and transparent decision-making by allowing the local community to decide who recieves grants.
@@
 
-Foreign donors already face many criticisms from grantees, including implementing models without taking local contexts into account and prioritizing projects that are more easily quantifiable and regime compatible.
+Foreign donors already face many criticisms from grantees, including implementing models without taking local contexts into account and prioritizing projects that are more easily quantifiable and regime compatible [@Bush:2015].
 In an environment of closing civic space, NGOs—especially those receiving grants from foreign donors—face even bigger risks.
@@
 Thus, NGOs not only seek to avoid working in communities repressed by the government, but also avoid working in areas where authorities use anti-NGO rhetoric.
-While donor responses have largely focused on navigating and adapting to repression, the use of such rhetoric raises broader questions about research and policy on the nonprofit sector: how is such rhetoric changing public attitudes and engagement with these groups?
-Given that anti-NGO repression and rhetoric can exacerbate burnout rates [@JoscelyneKnuckeySatterthwaite:2015], what are organizations doing to support human rights advocates, and what kinds of interventions can help support activists and professionals?
+While donor responses have largely focused on navigating and adapting to repression, the use of such rhetoric raises broader questions about research and policy in the nonprofit sector: how is such rhetoric changing public attitudes and engagement with these groups?
+Given that anti-NGO repression and rhetoric can exacerbate burnout rates [@JoscelyneKnuckeySatterthwaite:2015], what are organizations doing to support rights advocates, and what kinds of interventions can help support activists and professionals?
 In response to these concerns, organizations such as Defend Defenders, Frontline Defenders, Peace Brigades International, Civil Rights Defenders, and Protection International provide physical security, legal and digital support, and mental health resources to activists at risk.
@@
 
-While examples of Kenya and Kyrgyzstan discussed above show that donors and local activists can successfully push back against anti-NGO laws, a more systematic approach is needed.
-To tackle this administrative repression, Western states need a more effective and coordinated response that engages not only aid agencies, donors, and international NGOs but also businesses, development communities, diplomatic actors, and local populations to address larger structural issues that exacerbate government crackdowns on civil society.
+While the examples of Kenya and Kyrgyzstan discussed above show that donors and local activists can successfully push back against anti-NGO laws, a more systematic approach is needed.
+To tackle this administrative repression, Western states need a more effective and coordinated response that engages not only aid agencies, donors, and international NGOs but also businesses, development communities, and local populations to address larger structural issues that exacerbate government crackdowns on civil society.
 Without such an effort, the repression of civil society organizations may be the canary in the coal mine that sets the groundwork for future democratic erosion.

### 2024-11-05 manuscript.qmd
Update number

@@
 Recent research finds that national and multilateral aid agencies are responsive to anti-NGO laws, though often in inconsistent ways [@ChaudhryHeiss:2022a].
-@ChristensenWeinstein:2013 show that aid from bilateral donors—or direct country-to-country aid—sees an average decrease of $25–\$50 million after recipient states pass laws that restrict the flow of foreign funds to NGOs.
+@ChristensenWeinstein:2013 show that aid from bilateral donors—or direct country-to-country aid—sees an average decrease of $25 million after recipient states pass laws that restrict the flow of foreign funds to NGOs.
 However, aid from multilateral agencies is not affected by these laws.

### 2024-11-05 manuscript.qmd
Set up nicer HTML and PDF output

@@
 subtitle: "Chapter for the *Handbook of Political Control*"
+date: 2024-11-04
 author:
@@
     wordcount-banner: true
+    shift-heading-level-by: 1
+    theme: 
+      - zephyr
+      - html/custom.scss
+    template-partials: 
+      - html/title-metadata.html
     include-in-header:
@@
           </style>
+    knitr:
+      opts_chunk:
+        dev: "ragg_png"
+        dpi: 300
+
+  # Nice PDF output
+  hikmah-pdf:
+    toc: false
+    # Custom fonts
+    mainfont: "Linux Libertine O"
+    mainfontoptions:
+      - "Numbers=Proportional"
+      - "Numbers=OldStyle"
+    sansfont: "Inter"
+    mathfont: "Libertinus Math"
+    # linestretch: 1.05
+
+    knitr:
+      opts_chunk:
+        dev: "cairo_pdf"
 params:
@@
     <strong>{{< words-sum body-ref >}} total words</strong>: {{< words-body >}} in the body • {{< words-ref >}} in the references
+language: 
+  title-block-published: "Latest version"
 ---
@@
 library(ggraph)
+
+update_geom_defaults("text", list(family = "Inter"))
+update_geom_defaults("label", list(family = "Inter"))
 ```

### 2024-11-05 README.qmd
Update readme

@@
 
-# manuscript_target <- targets::tar_read(manuscript)
+yaml_metadata <- rmarkdown::yaml_front_matter("manuscript.qmd")
+title <- yaml_metadata$title
+# abstract <- yaml_metadata$abstract
+abstract <- "Given the increase in administrative and legal repression of civil society, how have their donors responded?
+How do these responses differ across foreign government and private donors?
+How have local philanthropists and NGOs adapted and responded to such repression?
+What challenges do civil society groups continue to face despite these responses?"
+```
 
-# yaml_metadata <- rmarkdown::yaml_front_matter(here::here("manuscript", "manuscript.qmd"))
-# title <- yaml_metadata$title
-# abstract <- yaml_metadata$abstract
-```
+# `r title`
+
+<!-- badges: start -->
+[![Preprint](https://img.shields.io/badge/Preprint-10.31235%2Fosf.io%2F9gmhp-blue)](https://doi.org/10.31235/osf.io/9gmhp) [![OSF DOI](https://img.shields.io/badge/OSF-10.17605%2FOSF.IO%2FR97Y3-blue)](https://doi.org/10.17605/OSF.IO/R97Y3) [![Code DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.12817616.svg)](https://doi.org/10.5281/zenodo.12817616)
+<!-- badges: end -->
+
+[Andrew Heiss](https://www.andrewheiss.com/) • Andrew Young School of Policy Studies • Georgia State University  
+[Suparna Chaudhry](https://www.suparnachaudhry.com/) • Lewis and Clark College
+
+---
+
+## Abstract
+
+`r abstract`
 

### 2024-11-05 README.qmd
Update Zenodo DOI

@@
 <!-- badges: start -->
-[![Preprint](https://img.shields.io/badge/Preprint-10.31235%2Fosf.io%2F9gmhp-blue)](https://doi.org/10.31235/osf.io/9gmhp) [![OSF DOI](https://img.shields.io/badge/OSF-10.17605%2FOSF.IO%2FR97Y3-blue)](https://doi.org/10.17605/OSF.IO/R97Y3) [![Code DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.12817616.svg)](https://doi.org/10.5281/zenodo.12817616)
+[![Preprint](https://img.shields.io/badge/Preprint-10.31235%2Fosf.io%2F9gmhp-blue)](https://doi.org/10.31235/osf.io/9gmhp) [![OSF DOI](https://img.shields.io/badge/OSF-10.17605%2FOSF.IO%2FR97Y3-blue)](https://doi.org/10.17605/OSF.IO/R97Y3) [![Code DOI](https://img.shields.io/badge/Code-10.5281%2Fzenodo.14038467-blue)](https://doi.org/10.5281/zenodo.14038467)
 <!-- badges: end -->

### 2024-11-05 deploy.sh
Add deploy script

@@ -1,0 +1,7 @@
+REMOTE_HOST="ath-cloud"
+REMOTE_DIR="~/sites/stats.andrewheiss.com/public_html/butterscotch-bat"
+REMOTE_DEST=$REMOTE_HOST:$REMOTE_DIR
+
+echo "Uploading new changes to remote server..."
+echo
+rsync -crvP --exclude '*_cache' --delete manuscript.html manuscript_files manuscript.pdf $REMOTE_DEST

### 2024-11-05 README.qmd
Add badges for HTML and PDF

--- 2024-11-05 README.qmd
@@
 <!-- badges: start -->
-[![Preprint](https://img.shields.io/badge/Preprint-10.31235%2Fosf.io%2F9gmhp-blue)](https://doi.org/10.31235/osf.io/9gmhp) [![OSF DOI](https://img.shields.io/badge/OSF-10.17605%2FOSF.IO%2FR97Y3-blue)](https://doi.org/10.17605/OSF.IO/R97Y3) [![Code DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.12817616.svg)](https://doi.org/10.5281/zenodo.12817616)
+[![HTML preprint](https://img.shields.io/badge/HTML%20preprint-FF851B)](https://stats.andrewheiss.com/butterscotch-bat/manuscript.html) [![PDF preprint](https://img.shields.io/badge/PDF%20preprint-3D9970)](https://stats.andrewheiss.com/butterscotch-bat/manuscript.pdf) [![SocArXiv Preprint](https://img.shields.io/badge/SocArXiv%20preprint-10.31235%2Fosf.io%2F9gmhp-blue)](https://doi.org/10.31235/osf.io/9gmhp) [![OSF DOI](https://img.shields.io/badge/OSF-10.17605%2FOSF.IO%2FR97Y3-blue)](https://doi.org/10.17605/OSF.IO/R97Y3) [![Code DOI](https://img.shields.io/badge/Code-10.5281%2Fzenodo.14038467-blue)](https://doi.org/10.5281/zenodo.14038467)
 <!-- badges: end -->

### 2024-11-05 index.qmd
Rename manuscript to index

--- 2024-11-05 manuscript.qmd
@@
 author:
-- name: Suparna Chaudhry
-  affiliation: Lewis and Clark College
-  orcid: 0000-0003-3506-7590
-  url: https://www.suparnachaudhry.com/
-  email: schaudhry@lclark.edu
 - name: Andrew Heiss
@@
     corresponding: true
+- name: Suparna Chaudhry
+  affiliation: Lewis and Clark College
+  orcid: 0000-0003-3506-7590
+  url: https://www.suparnachaudhry.com/
+  email: schaudhry@lclark.edu
 bibliography: references.bib

### 2024-11-05 index.qmd
Convert to project

--- 2024-11-05 manuscript.qmd
@@
 subtitle: "Chapter for the *Handbook of Political Control*"
+date: 2024-11-04
 author:
-- name: Suparna Chaudhry
-  affiliation: Lewis and Clark College
-  orcid: 0000-0003-3506-7590
-  url: https://www.suparnachaudhry.com/
-  email: schaudhry@lclark.edu
 - name: Andrew Heiss
@@
     corresponding: true
+- name: Suparna Chaudhry
+  affiliation: Lewis and Clark College
+  orcid: 0000-0003-3506-7590
+  url: https://www.suparnachaudhry.com/
+  email: schaudhry@lclark.edu
 bibliography: references.bib
 csl: csl/chicago-author-date.csl
-toc: true
-format:
-  wordcount-html: 
-    wordcount-banner: true
-    include-in-header:
-      - text: |
-          <style type="text/css">
-          h2.hide {
-            display: none;
-          }
-          </style>
 params:

### 2024-11-05 deploy.sh
Deploy output folder

@@ -4,4 +4,4 @@ REMOTE_DEST=$REMOTE_HOST:$REMOTE_DIR
 
 echo "Uploading new changes to remote server..."
 echo
-rsync -crvP --exclude '*_cache' --delete manuscript.html manuscript_files manuscript.pdf $REMOTE_DEST
+rsync -crvP --exclude '*_cache' --delete _output/ $REMOTE_DEST

### 2024-11-05 index.qmd
Add some links and tweaks

@@
   email: schaudhry@lclark.edu
+
 bibliography: references.bib
 csl: csl/chicago-author-date.csl
+
+doi: 10.31235/osf.io/9gmhp
+doi-title: SocArXiv preprint
+
 params:

### 2024-11-05 README.qmd
Add linebreak after preprint badges
Change name of PDF

--- 2024-11-05 README.qmd
@@
 <!-- badges: start -->
-[![Preprint](https://img.shields.io/badge/Preprint-10.31235%2Fosf.io%2F9gmhp-blue)](https://doi.org/10.31235/osf.io/9gmhp) [![OSF DOI](https://img.shields.io/badge/OSF-10.17605%2FOSF.IO%2FR97Y3-blue)](https://doi.org/10.17605/OSF.IO/R97Y3) [![Code DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.12817616.svg)](https://doi.org/10.5281/zenodo.12817616)
+[![HTML preprint](https://img.shields.io/badge/HTML%20preprint-FF851B)](https://stats.andrewheiss.com/butterscotch-bat/) [![PDF preprint](https://img.shields.io/badge/PDF%20preprint-3D9970)](https://stats.andrewheiss.com/butterscotch-bat/heiss-chaudhry-donor-responses-chapter.pdf) [![SocArXiv Preprint](https://img.shields.io/badge/SocArXiv%20preprint-10.31235%2Fosf.io%2F9gmhp-blue)](https://doi.org/10.31235/osf.io/9gmhp)
+[![OSF DOI](https://img.shields.io/badge/OSF-10.17605%2FOSF.IO%2FR97Y3-blue)](https://doi.org/10.17605/OSF.IO/R97Y3) [![Code DOI](https://img.shields.io/badge/Code-10.5281%2Fzenodo.14038467-blue)](https://doi.org/10.5281/zenodo.14038467)
 <!-- badges: end -->

### 2024-11-05 README.qmd
oops, get metadata from actual manuscript

@@
 
-yaml_metadata <- rmarkdown::yaml_front_matter("manuscript.qmd")
+yaml_metadata <- rmarkdown::yaml_front_matter("index.qmd")
 title <- yaml_metadata$title

### 2024-11-05 README.qmd
Add abstract

@@
 title <- yaml_metadata$title
-# abstract <- yaml_metadata$abstract
-abstract <- "Given the increase in administrative and legal repression of civil society, how have their donors responded?
-How do these responses differ across foreign government and private donors?
-How have local philanthropists and NGOs adapted and responded to such repression?
-What challenges do civil society groups continue to face despite these responses?"
+abstract <- yaml_metadata$abstract
 ```

### 2024-11-05 index.qmd
Add abstract

@@
   email: schaudhry@lclark.edu
+
+abstract: |
+  This chapter examines donor responses to the global crackdown on civil society, where governments use legal measures to repress NGOs.
+Donors, including bilateral and multilateral agencies, respond variably; bilateral aid often decreases following anti-NGO laws, while multilateral aid remains stable.
+Private donors' responses depend on factors like social trust and NGO transparency.
+NGOs adapt by diversifying funding, forming affiliates, or shifting to community-based support.
+As the space for civil society continues to shrink, coordinated international efforts are essential to counteract these repressive trends and uphold civil society's role in democratic governance.
 

### 2025-01-23 index.qmd
Include hidden introduction heading for the TOC
Add revisions

@@
 subtitle: "Chapter for the *Handbook of Political Control*"
-date: 2024-11-04
+date: 2025-01-23
 author:
@@
 
-::: {.content-visible when-format="wordcount-html"}
+::: {.content-visible when-format="html"}
 # Introduction {.hide}
@@
 As of 2023, 87% of the world's population now lives in countries where the space available to civil society is closed, repressed, or obstructed [@CIVICUSMonitor].
+This intersection of administrative and legal institutions (layer 3) being used to control the activities and existence of civil society organizations (layer 4) is an example of how Earl and Braithwaite's [-@EarlBraithwaite:2022] layered model helps us consider the tools states use to control dialogue and mobilization among their populations.
 
-The growth in states using law to repress can be attributed to three main factors.
-First, administrative or legal crackdowns have many advantages over the use of violence.
+The growth in states using law to repress civil society organizations can be attributed to three main factors.
+First, administrative or legal crackdown has many advantages over the use of violence.
 It allows governments to maintain a democratic facade and limit the domestic backlash because citizens tend to view legal crackdown as regulation rather than repression [@Chaudhry:2022].
@@
 Conversely, continuing aid might signal tolerance for rights violations, weakening the credibility of the donor's human rights agenda.
-Prior research finds conflicting relationships between human rights abuses and foreign aid, variously concluding that donor agencies punish repressive states by reducing or withdrawing aid [@LebovicVoeten:2009], increasing aid to repressive states as a way to counter state violence [@Nielsen:2013], or making no measurable change in aid allocations [@Neumayer:2003; @Neumayer:2003b].
+Prior research finds conflicting relationships between human rights abuses and foreign aid, variously concluding that donor agencies punish repressive states by reducing or withdrawing aid [@LebovicVoeten:2009], increasing aid to repressive states to counter state violence [@Nielsen:2013], or making no measurable change in aid allocations [@Neumayer:2003; @Neumayer:2003b].
 
 One reason for such mixed results is that the decision to withdraw or suspend aid in response to human rights abuses (including civil society crackdown) is often seen as a binary choice driven by self-interested motivations: a donor country that sees human rights abuses in a recipient state will consider its own foreign policy interests before deciding to either continue or curtail aid.
-In reality, however, the decision is far more complex and involves a wide range of possible motivations and responses.
+However, the decision is far more complex and involves a wide range of possible motivations and responses.
 @CheesemanSwedlundOBrien-Udry:2024 offer a stylized decision tree of possible choices following recipient-country human rights violations—we provide a simplified version of their framework in @fig-agency-response-tree.
@@
 
-Research on aid conditionality and agency responses to human rights abuses tends to look at more overt and violent forms of political repression.
-The effects of more subtle administrative crackdown on aid, including the passage of anti-NGO laws and the constricting of civic space, are less explored.
+Research on aid conditionality and agency responses to human rights abuses tends to look at more overt and violent forms of political repression, or higher layers in the @EarlBraithwaite:2022 framework.
+The effects of more subtle administrative crackdown on aid—including the passage of anti-NGO laws, the constricting of civic space, and other layer 3 institutions—are less explored.
 Civil society plays a central role in the distribution and implementation of international aid.
@@
 Human rights abuses in recipient countries will shape the responses of donor countries depending on the saliency of the threat to aid provision: political repression that threatens business interests and investments will be more salient to right-leaning governments, while anti-NGO civil society administrative crackdown will likely have a stronger effect on left-leaning governments' responses, since domestic NGOs are central to donor countries' aid distribution strategies.
+This dynamic indicates ways that the @EarlBraithwaite:2022 framework might be expanded to work between states—layer 3 institutions like civil society regulations in donor countries can affect perceptions of layer 3 institutions in recipient countries, and vice versa.
 
@@
 This model may work well for civic groups working on relatively tame causes or exclusively development-focused goals, but more contentious organizations facing government crackdown may need flexible funding that does not bring immediate results or does not have easily measurable goals [@Herrold:2020].
+
+This watering down of NGO programming illustrates interesting interplay between the different layers of Earl and Braithwaite's [-@EarlBraithwaite:2022] model of repression.
+Both de jure anti-NGO laws and the de facto civil society regulatory environment represent layer 3 institutions designed to limit participation in institutional politics.
+These limits cause civil society actors at layer 4 to shift to safer strategies and curtail their programming, which in turn can lead to worse consequences downstream.
+Formal and informal limits on civil society could be one mechanism behind the fact that foreign democracy aid increases the probability of civil war onset in states with weaker and less institutionalized civil society [@BraithwaiteLicht:2020].
+Amid violent level 7 conflict, civil society restrictions may prevent NGOs from intervening effectively to address humanitarian challenges arising from violent conflict.
+For instance, Executive Order 13224—issued by George W.
+Bush in 2001—banned all financial transactions with any individual or organization on a list of Specially Designated Global Terrorists [@eo13224journal].
+However, many humanitarian aid organizations (layer 4 institutions), work in environments where even the smallest circumstantial evidence could be grounds for designating these NGOs as having provided material support or resources for terrorism [@Bhungalia:2023].
+Thus, many nonprofits may choose to avoid working in conflict zones.
+For example, after the US listed Al Shabaab in Somalia as a terrorist group, it resulted in a 88% reduction in civil society aid to Somalia over the next two years, with devastating consequences for Somalis [@Hayes:2017].
+
+Other layers in this model of repression also point toward future research on donor responses to NGO restrictions.
+@BraithwaiteLicht:2020 find that states that receive foreign democracy assistance have a lower probability of internal conflict when civil society is more institutionalized.
+It is likely that broader social institutionalization (e.g., strong formal social institutions [level 5] and robust protest mobilization [level 6]) protects against more violent repression.
+We do not yet know how these other levels of institutionalization interact with level 3 civil society restriction, nor how official aid donors might condition their responses to repression.
+For instance, are donor states less likely to tame their programming when there are popular protests or powerful labor unions that can supplement resistance to government action?
 
@@
 Building on this work, @ChaudhryDotsonHeiss:2024 examine how organizational characteristics (e.g., issue area, funding sources) and practices (e.g., financial transparency) interact with structural concerns like legal crackdown abroad.
-They find that, all else equal, legal crackdown on its own causes donors to reduce their support for nonprofits—individuals are 5–10 percentage points less likely to donate to an NGO that is criticized or restricted, compared to an organization that enjoys a friendly relationship with its host government.
+They find that, all else equal, legal crackdown on its own causes donors to reduce their support for nonprofits—individuals are less likely to donate to an NGO that is criticized or restricted, compared to an organization that enjoys a friendly relationship with its host government.
 However, this reduction in support can be offset by organizational characteristics.
-Financial transparency and accountability protect each increase the probability of donations by nine percentage points under the worst conditions of legal crackdown.
+Financial transparency and accountability each increase the probability of donations under the worst conditions of legal crackdown.
 Individual donors thus typically reduce their support when seeing legal crackdown, but change their response and increase their support when seeing other signs that the organization follows best practices and is more deserving.
@@
 Senate Committee on Foreign Relations hearing maintained that "[n]either remittances nor philanthropy can replace the expertise, scale, or agenda setting capacity of American foreign policy assistance" [@Runde:2017].
-Thus, despite promising new research on how to raise funds from private sources for besieged NGOs abroad, official aid donors also need to mount a coherent response to the shrinking civic space.
+Thus, despite promising new research on how to raise funds from private sources for besieged NGOs abroad, official aid donors also need to mount a coherent response to the shrinking civic space, in which the actors comprising layer 4 of the @EarlBraithwaite:2022 model are challenged by restrictions from layer 3 institutions.
 
@@
 To tackle this administrative repression, Western states need a more effective and coordinated response that engages not only aid agencies, donors, and international NGOs but also businesses, development communities, and local populations to address larger structural issues that exacerbate government crackdowns on civil society.
+This chapter demonstrates that the use of layer 3 controls---such as law and administrative repression---does not just have an impact on layer 4 actors such as civil society institutions.
+They also have far-reaching repercussions on efforts to address and resolve violent conflict (layer 7) and support for vulnerable communities, including minorities, queer groups, women, refugees, and asylum seekers (layer 2), who often depend on specialized nonprofit groups for services and legal advice that may cease to exist in the face of legal restrictions and anti-NGO rhetoric.
 Without such an effort, the repression of civil society organizations may be the canary in the coal mine that sets the groundwork for future democratic erosion.

### 2025-03-16 index.qmd
Add keywords

@@
 subtitle: "Chapter for the *Handbook of Political Control*"
-date: 2025-01-23
+date: 2025-03-16
 author:
@@
 As the space for civil society continues to shrink, coordinated international efforts are essential to counteract these repressive trends and uphold civil society's role in democratic governance.
+
+keywords: 
+  - foreign aid
+  - NGOs
+  - NGO crackdown
+  - civil society repression
+  - international philanthropy
+  - charitable giving
 
@@
 Formal and informal limits on civil society could be one mechanism behind the fact that foreign democracy aid increases the probability of civil war onset in states with weaker and less institutionalized civil society [@BraithwaiteLicht:2020].
-Amid violent level 7 conflict, civil society restrictions may prevent NGOs from intervening effectively to address humanitarian challenges arising from violent conflict.
+Amid violent layer 7 conflict, civil society restrictions may prevent NGOs from intervening effectively to address humanitarian challenges arising from violent conflict.
 For instance, Executive Order 13224—issued by George W.
@@
 @BraithwaiteLicht:2020 find that states that receive foreign democracy assistance have a lower probability of internal conflict when civil society is more institutionalized.
-It is likely that broader social institutionalization (e.g., strong formal social institutions [level 5] and robust protest mobilization [level 6]) protects against more violent repression.
-We do not yet know how these other levels of institutionalization interact with level 3 civil society restriction, nor how official aid donors might condition their responses to repression.
+It is likely that broader social institutionalization (e.g., strong formal social institutions [layer 5] and robust protest mobilization [layer 6]) protects against more violent repression.
+We do not yet know how these other layers of institutionalization interact with layer 3 civil society restriction, nor how official aid donors might condition their responses to repression.
 For instance, are donor states less likely to tame their programming when there are popular protests or powerful labor unions that can supplement resistance to government action?
@@
 Without such an effort, the repression of civil society organizations may be the canary in the coal mine that sets the groundwork for future democratic erosion.
+Recent developments within Western democracies---including the second Trump administration's attacks on both the USAID and the U.S.
+nonprofit sector and European donors' increasing withdrawal from external democracy support---raise new questions about how civic groups and private donors and foundations can respond to a changing international landscape.
 

### 2025-04-28 index.qmd
Add alt text

@@
 subtitle: "Chapter for the *Handbook of Political Control*"
-date: 2025-03-16
+date: 2025-04-28
 author:
@@
 #| fig-cap: Possible donor- and recipient-country responses to human rights violations
+#| fig-alt: |
+#|   A flowchart showing possible reactions to human rights oppression.
+#|   Recipient governments violate human rights. Donor governments can respond by
+#|   doing nothing, threatening the recipient government, withdrawing or suspending
+#|   aid, or shifting aid. Recipient governments can counter-respond by doing
+#|   nothing, modifying policies to improve relationships, or pressuring donors
+#|   to back down.
 #| include: true
@@
 #| out-width: 80%
+
+# fig-alt-short: "A flowchart showing possible reactions to oppression: recipient governments violate human rights, donors respond, and recipients counter-respond."
 
@@
 Donor countries have a range of possible responses when recipient countries engage in administrative crackdown.
-Recent research finds that national and multilateral aid agencies are responsive to anti-NGO laws, though often in inconsistent ways [@ChaudhryHeiss:2022a].
+Recent research finds that national and multilateral aid agencies are responsive to anti-NGO laws, though often in inconsistent ways [@ChaudhryHeiss:2022].
 @ChristensenWeinstein:2013 show that aid from bilateral donors—or direct country-to-country aid—sees an average decrease of $25 million after recipient states pass laws that restrict the flow of foreign funds to NGOs.
@@
 Donor countries respond differently to these types of oppression.
-@ChaudhryHeiss:2024 find that bilateral donors are quite responsive to barriers to advocacy—on average, total foreign aid decreases by 30–40% (or \$100 in a typical country) following the passage of an advocacy-focused NGO law [see also @ChaudhryHeiss:2022a].
+@ChaudhryHeiss:2024 find that bilateral donors are quite responsive to barriers to advocacy—on average, total foreign aid decreases by 30–40% (or \$100 in a typical country) following the passage of an advocacy-focused NGO law [see also @ChaudhryHeiss:2022].
 @RightSpringmanWibbels:2024 find similar and more nuanced trends in donor responses: donors committed to political advocacy (i.e.
@@
 #| fig-cap: Possible individual responses to civil society restrictions in recipient countries
+#| fig-alt: |
+#|   A flowchart showing possible individual and NGO reactions to civil society
+#|   restrictions. NGO host governments restrict civil society. Individual private
+#|   donors can respond by doing nothing, increasing their donations, or decreasing 
+#|   their donations. NGOs then respond in an unknown way.
 #| include: true
@@
 #| out-width: 80%
+
+# fig-alt-short: "A flowchart showing possible individual and NGO reactions to civil society restrictions: governments limit civil society, individual donors respond."
 
